### 2019-05-04 doc/paper.tex
feito registro dos andamentos p finalizar artigo

% ****** Start of file aipsamp.tex ******
%
%   This file is part of the AIP files in the AIP distribution for REVTeX 4.
%   Version 4.1 of REVTeX, October 2009
%
%   Copyright (c) 2009 American Institute of Physics.

% Use this file as a source of example code for your aip document.
% Use the file aiptemplate.tex as a template for your document.
\documentclass[%
 aip,
 jmp,%
 amsmath,amssymb,
%preprint,%
 reprint,%
%author-year,%
%author-numerical,%
]{revtex4-1}

\usepackage{graphicx}% Include figure files
\usepackage{grffile}
\usepackage{dcolumn}% Align table columns on decimal point
\usepackage{bm}% bold math
%\usepackage[mathlines]{lineno}% Enable numbering of text and display math
%\linenumbers\relax % Commence numbering lines
\usepackage{multirow}
\usepackage{color} % for the notes
\usepackage{etex}
\reserveinserts{58}
\usepackage{morefloats}

\maxdeadcycles=1000

\begin{document}

\preprint{XXXXX (preprint)}

%\title[Evolution of interaction networks]{On the evolution of interaction networks: primitive typology of vertex, prominence of measures and activity statistics}% Force line breaks with \\
%\title[Evolution of interaction networks]{On the evolution of interaction networks: a primitive typology of vertex}% Force line breaks with \\
\title[Versinus visualization of graphs]{Versinus: a visualization method for graphs in evolution}% Force line breaks with \\

\author{Renato Fabbri}%
 \homepage{http://ifsc.usp.br/~fabbri/}
 \email{fabbri@usp.br}
  \affiliation{ 
S\~ao Carlos Institute of Physics, University of S\~ao Paulo (IFSC/USP)%\\This line break forced with \textbackslash\textbackslash
}%
%
%\author{Camilo Akimushkin Valencia}
%  \homepage{http://ifsc.usp.br/}
%  \email{akimushkin@ifsc.usp.br}
%  \altaffiliation[Also at ]{IFSC-USP}%Lines break automatically or can be forced with \\
%
%
%\author{Vilson Vieira da Silva Junior}
%  \homepage{http://automata.cc/}
%  \email{vilson@void.cc}
%  \altaffiliation[Also at ]{IFSC-USP}%Lines break automatically or can be forced with \\
%
%\author{Ricardo Fabbri}
%  \homepage{http://www.lems.brown.edu/~rfabbri/}
%  \email{rfabbri@iprj.uerj.br}
% \altaffiliation{
%Instituto Polit\'ecnico, Universidade Estadual do Rio de Janeiro (IPRJ)
%}%Lines break automatically or can be forced with \\
%
%\author{Deborah Christina Antunes}
%  \homepage{http://lattes.cnpq.br/1065956470701739}
%  \email{deborahantunes@gmail.com}
%  \altaffiliation{
%Curso de Psicologia, Universidade Federal do Cer\'a (UFC)
%}%Lines break automatically or can be forced with \\
%
%\author{Marilia Mello Pisani}
%  \homepage{http://lattes.cnpq.br/6738980149860322}
%  \email{marilia.m.pisani@gmail.com}
% \altaffiliation{
%Centro de Ciências Naturais e Humanas, Universidade Federal do ABC (CCNH/UFABC)
%}%Lines break automatically or can be forced with \\
%
%
%\author{Luciano da Fontoura Costa}
%  \homepage{http://cyvision.ifsc.usp.br/~luciano/}
%  \email{ldfcosta@gmail.com}
%  \altaffiliation[Also at ]{IFSC-USP}%Lines break automatically or can be forced with \\
%
%\author{Osvaldo N. Oliveira Jr.}
%  \homepage{www.polimeros.ifsc.usp.br/professors/professor.php?id=4}
%  \email{chu@ifsc.usp.br}
% \altaffiliation[Also at ]{IFSC-USP}%Lines break automatically or can be forced with \\
%

\date{\today}% It is always \today, today,
             %  but any date may be explicitly specified

\begin{abstract}
This article presents a novel visualization approach for dynamic graphs, the versinus method,
specially useful for real world networks exhibiting free-scale properties. With a simple and fixed layout, and a small set of visual markups, the method has been useful for understanding network dynamics. Local community often suggests that it be reported, which motivated this article. Online resources deliver videos and computer scripts for rendering new animations. This article has a concise description of the method.
\end{abstract}

\pacs{01.50.Fr,89.75.Fb,05.65.+b,89.65.-s}% PACS, the Physics and Astronomy
\keywords{complex networks, visualization, social network analysis, pattern recognition, statistics}%Use showkeys class option if keyword
\maketitle

%\begin{quotation}
%% `The reason for the persistent plausibility of the typological approach, however, is not a static biological one, but just the opposite: dynamic and social. The fact that human society has been up to now divided into classes affects more than the external relations of men. The marks of social repression are left within the individual soul.' 
%% 
%`The conception of personality structure is the
%best safeguard against the
%inclination to attribute persistent trends in the
%individual to something
%"innate" or "basic" or "racial" within him. The
%Nazi allegation that natural, biological traits decide the total being of a person
%would not have been such
%a successful political device
%had it not been possible to point to numerous
%instances of relative fixity in human behavior and to
%challenge those who
%thought to explain them on any basis other than a biological one.'
%
%\emph{- Adorno et al, 1969, p. 747}
%\end{quotation}

\section{\label{sec:into}Introduction}
Versinus is a visualization method for dynamic graphs based on experimental observations.
This method receives this dedicated article by recurrence of the suggestion, by fellow researchers,
to write it.

In visualizing a network, the method consists of creating an animation,
of a fixed-size message sliding window (e.g. 400 messages), and 
partitioning the network in two fixed-layout segments:
a sinusoid for the most connected vertexes
and a straight line for the less connected.
A vertex holds the same position throughout the animation. Also,
visual indication of properties - such as color, height and width,
and rank of vertex with degree criteria - plays a central role.
This work differs from the few works on the visualization of dynamic
graphs because it is a simple report on a method that has been useful,
was developed by practical needs, is the result of experimentations,
and has a number of criteria that guided its development~\cite{Viz1,Viz2,Viz3}.

Next section further exposes the context in which versinus is being useful,
and the characteristics of the systems in which it was applied.
Section~\ref{vmethod} describes the method itself. Section~\ref{res} is dedicated
to most important results and discussions. Conclusions and further work
are in Section~\ref{conc}. Appendix exposes online resources.

\section{Context and materials}\label{cmaterials}
In 2013, while doing complex networks research, within a physics department,
to visualize networks undergoing its evolution along time was a necessity.
This posed some challenges (e.g. proper parametrization) to which two
visualization methods were developed. The first is a web gadget, accessible via
usual web browsers, that could render image galleries with
measures and the node-edge structure~\cite{galGMANE, appGMANE}.
The second method 
is based on rendering animations with specific characteristics.
This second method has been named \emph{versinus} for the purpose of this article.

\subsection{GMANE email lists interaction networks}\label{intNet}
The networks undergoing analysis were email interaction networks.
Messages were obtained from the GMANE public database~\cite{GMANE,GMANE2,GMANEwikipedia}. For this goal, just three metadata fields of each email message is needed: the sender, the ID of the message and the ID of the message it is a response to, if any. Each sender is a vertex in the network. If a message is a response, it yields an edge from the author of the original message to current message author. If the edge is already there, the weight is increased by one (an edge is created with weight one). By inspecting all messages in a set, the interaction network given by those messages can be attained. This interaction network is understood as an ``information network'', with edges along the information flow, as the response is an evidence that the responder absorbed the information given by the original sender. If the edges are inverted, it is usual to call it a ``status network'', with the responder "giving status" to the original sender, as he considered the message worthy enough to respond it.

\subsection{Sliding message window dynamic interaction networks}\label{sec:sliding}
Consider $\Delta$ a fixed number of consecutive messages (e.g. $\Delta=400$). Consider sets $s_{i}^{i+\Delta}$ of $\Delta$ consecutive email messages, as they are registered along time. A Sequence $S^{\Delta,M}$ of such sets, with the first message positioned in each the $M$ considered messages (e.g. $M=20000$), can be written as:

\begin{equation}
S^{\Delta,M}=\{s_i^{i+\Delta}\}_{i=0}^{M-\Delta}
\end{equation}

Each set $s_i$ yields an interaction network, as described in Section~\ref{intNet}. Each of these sets exhibits stable properties, while each participant exhibits a wide variation of characteristics along $S$. To understand the mechanisms of this compatibility (unstable vertexes, stable network) led to experimenting a series of layouts and visualization techniques, from which versinus emerged.

\section{The versinus visualization method}\label{vmethod}
Versinus is a simple visualization method. Network vertexes are roughly split to usual $80\%$ periphery, $15\%$ intermediary and $5\%$ of hubs. Hubs are laid on the first half of a sinusoid. Intermediary on the second. Peripheral sector is laid on a straight line. This configuration can be improved in various forms, to which Section~\ref{sec:ref} is dedicated. Figure~\ref{fig:versinus} has an image of such a layout. The position of vertex, fixed, is defined by the overall structure, i.e. with respect to all $M$ messages (see Section~\ref{sec:sliding}). Numbers with individual measures for each vertex blink periodically.

\begin{figure}[h!]
    \begin{center}
        \includegraphics[scale=.25]{figs/versinus_}
        \caption{The versinus visualization method in use. 5\% of The most connected vertexes (hubs) are on the left period of the sinusoid. 15\% of the most connected remaining vertexes are on the right period. 80\% of the least connected vertexes are on the straight line, above the sinusoidal shape. White dots with numbers keep track of node position in the overall degree ordering. Measures blink periodically on the vertex it is related to.}
        \label{fig:versinus}
    \end{center}
\end{figure}

\section{Results and discussion}\label{res}
Results from versinus is divided in two groups:
observations on features that made it useful for the task,
and the network structure properties it made possible to grasp.

\subsection{Understanding of useful visualization features for dynamic networks}

On the numerous insights related to versinus, a few of them seem more fundamental, or plain useful. This is an attempt to present them in an importance-first order:
\begin{enumerate}
    \item Vertexes need to remain static. Even if they move smoothly, one notices solely transient artifacts if nodes are moving around.
    \item Very connected sectors (hubs and intermediary) need to be laid in a curve, otherwise the edges enclose each other and reasoning about the network becomes harsh.
    \item Height and width of a vertex are very informative, specially if measures mapped to them have a strong relation, such as out-degree (mapped to height in versinus) and in-degree (mapped to width).
    \item Coloring nodes was also informative, although less than height and weight, as these are related to measure differences more directly.
    \item An ordering of nodes, related to their fixed position, is very useful. Among all tests, ordering of vertexes by degree was considered the most informative order, which led to the hub, intermediary and peripheral sectioning. As node position in layout is fixed, its ordering is done with respect to the overall structure, i.e. considering all $M$ messages.
    \item  Numbering these positions with respect to the order of the vertexes in the larger structure (all $M$ messages) is useful in understanding how much a vertex holds his position in his scale and on a larger window size.
\end{enumerate}

Many other insights were given by versinus, such as possible visualization tools, other kind of convenient layouts and glyph elaborations. These received dedicated attention in Section~\ref{sec:ref}.
Even so, those insights, numbered above, were incorporated to versinus as the result of tests which presented clear benefits within the best settings achieved.

\subsection{Understanding of network properties through versinus}
A number of hypothesis were drawn about networks, for which versinus
was designed. Another number of hypothesis were driven from versinus use itself.

\subsubsection{Hypothesis evidenced by versinus}
As suggested by Palla, Barab\'asi and Vicsek~\cite{barabasiEvo}, stability of participant activity in social networks is more incident in smaller networks. In accordance with this result, all hubs have intermittent activity in the settings analyzed, except for the email list with the smallest number of participants (the Metareciclagem email list). The intermitence of hubs itself was one of the top hypothesis which motivated versinus development. The stability of the structure~\cite{evoSN} required further verification of the instability of each participant activity.

\subsubsection{Hypothesis drawn from versinus use}
There seem to be modes of operation of the network. As an example, the intermediary is observed to communicate mostly to hubs or to communicate substantially with the peripheral vertex. Another hypothesis is that in and out degree discrepancies in each vertex are important for network characterization. This was confirmed by a recent work by the present author~\cite{evoSN}. Other hypothesis, such as discrepancies in authority and connectivity of vertex, are numerous but need further research to be valuable.

\section{Conclusions and further work}\label{conc}
Versinus has been calling the attention of fellow
researchers for its simplicity and utility. This concise
exposition of the method is aimed at giving further
consequence for this enthusiasm. Results are satisfactory
in the domain of interest and applications of the method
to visualize other domain dynamic networks should come in near future.

\subsection{Refinements on versinus}\label{sec:ref}
By far, the most proficuous facet of versinus is its ability
to bring insights about how to enhance its layout and use.
Besides the technological proposal of making a tool for using versinus
in real-time, there are some ideas about the layout and visual guides itself.

To further enable visualization of hubs and intermediary vertex,
the sinusoid can have many periods 
with a decaying frequency.
The upper straight line can also have an oscillating trace.
The two halves of the sinusoidal period can be moved independently.
The waveform need not to be a sinusoid.
One can think of many ways to make more informative glyphs. Also, visual and auditory signals for specific occurrences can be interesting (e.g. when a new vertex appears, when one vanishes, when an ordering of vertexes changes).
Measures of each vertex can be exposed with a vertical displacement, to enable multiple measures, to avoid the necessity to blink the numbers and to keep network visualization free from occlusion.

\subsection{Other strategies envisioned}\label{sec:other}
Working with versinus has suggested other kinds of layout for vertexes, specially other geometric figures and iterative force-based methods for positioning vertex in a fixed layout. Matrix visualization of graphs and recent approaches~\cite{Viz1} has been foreseen as support to versinus.

\appendix

\section{Online videos}
There are video playlists online for use by the research team. Authors
choose~\cite{animacoes} as an example. Other playlists can be found by inspecting the user.

\section{Using online scripts for rendering versinus animations}
For making the animations and the analysis of the networks, Python scripts are online in a public repository~\cite{scriptsFim}.

\vspace{5mm}
\nocite{*}
\bibliography{paper}% Produces the bibliography via BibTeX.

\end{document}
%
% ****** End of file aipsamp.tex ******

### 2019-05-04 legacy/sistemaMinimo_5000Foo.py

@@ -1,0 +1,567 @@
+#-*- coding: utf8 -*-
+from __future__ import division
+import numpy as n, pylab as p, networkx as x, random as r, collections as c, string
+
+__doc__="""Este arquivo possui a classe Sistem, base para todas as animações
+
+G=x.read_gml("1-400cpp.gml") # digrafo com peso
+S=Sistem(G)
+S.draw("grafo1.png")
+S.add_msgs([msg1,msg2...])
+S.rm_msgs([msg1,msg2...])
+S.order()
+S.draw("grafo1.png")
+
+# nadds[i] tem as msgs adicionadas para o frame i+1.
+nadds=[[msgs..,..,],[msgs..],[]]
+# nrms[i] tem as msgs removidas para o frame i+1.
+nrms=[[msgs..,..,],[msgs..],[]]
+i=0
+for a,r in zip(nadds,nrms):
+    S.add_msgs(a)
+    S.rm_msgs(b)
+    S.order()
+    S.draw("grafo%i.png"%(i,))
+    i+=1
+
+0) fazer classe com parametros minimos. Fazer com que plote pelo pygraphviz sem layout.
+1) fazer classe com parametros minimos. Fazer com que plote pelo pygraphviz com layout pre-determinado.
+"""
+
+class Sistem:
+    def __init__(self,G="lau300_500.gml",positionMethod="sinusoid",rankCriteria="degree",fixedSize=40):
+        """G Digrafo do networkx ou gml para
+        
+        positionMethod: "random" ou "sinusoid"
+        rankCriteria: "alphabetic" ou "degree" """
+        self.rank=None
+        self.positions=None
+        self.ecolors=[]
+        if type(G) == type("string"):
+            G=x.read_gml(G) # DiGraph
+        self.g=G
+        self.positionMethod=positionMethod
+        self.rankCriteria=rankCriteria
+        self.fixedSize=fixedSize
+        N=1. # aresta do quadrado considerado
+        # ranks nodes in an exact order
+        rank=self.rankNodes() # implement
+        # gets (x,y) pairs for each ranked node:
+        positions=self.positionRanks() 
+        ruido=self.computeNoise()
+
+    def update(self):
+        self.rankNodes()
+        self.positionRanks()
+
+    def addNode(self,nodeName):
+        self.g.add_node(nodeName,weight=1.0)
+        self.update()
+    def addEdge(self,pre,pos):
+        self.g.add_edge(pre,pos,weight=1.0)
+        self.update()
+
+    def rankNodes(self,force=False):
[…]

### 2019-05-08 doc/paper.tex
inserido novo template para novo escrito

@@
-% ****** Start of file aipsamp.tex ******
+% This is samplepaper.tex, a sample chapter demonstrating the
+% LLNCS macro package for Springer Computer Science proceedings;
+% Version 2.20 of 2017/10/04
 %
-%   This file is part of the AIP files in the AIP distribution for REVTeX 4.
-%   Version 4.1 of REVTeX, October 2009
+\documentclass[runningheads]{llncs}
 %
-%   Copyright (c) 2009 American Institute of Physics.
-
-% Use this file as a source of example code for your aip document.
-% Use the file aiptemplate.tex as a template for your document.
-\documentclass[%
- aip,
- jmp,%
- amsmath,amssymb,
-%preprint,%
- reprint,%
-%author-year,%
-%author-numerical,%
-]{revtex4-1}
-
-\usepackage{graphicx}% Include figure files
-\usepackage{grffile}
-\usepackage{dcolumn}% Align table columns on decimal point
-\usepackage{bm}% bold math
-%\usepackage[mathlines]{lineno}% Enable numbering of text and display math
-%\linenumbers\relax % Commence numbering lines
-\usepackage{multirow}
-\usepackage{color} % for the notes
-\usepackage{etex}
-\reserveinserts{58}
-\usepackage{morefloats}
-
-\maxdeadcycles=1000
+\usepackage{graphicx}
+% Used for displaying a sample figure. If possible, figure files should
+% be included in EPS format.
+%
+% If you use the hyperref package, please uncomment the following line
+% to display URLs in blue roman font according to Springer's eBook style:
+% \renewcommand\UrlFont{\color{blue}\rmfamily}
 
 \begin{document}
+%
+\title{Contribution Title\thanks{Supported by organization x.}}
+%
+%\titlerunning{Abbreviated paper title}
+% If the paper title is too long for the running head, you can set
+% an abbreviated paper title here
+%
+\author{First Author\inst{1}\orcidID{0000-1111-2222-3333} \and
+Second Author\inst{2,3}\orcidID{1111-2222-3333-4444} \and
+Third Author\inst{3}\orcidID{2222--3333-4444-5555}}
+%
+\authorrunning{F. Author et al.}
+% First names are abbreviated in the running head.
+% If there are more than two authors, 'et al.' is used.
+%
+\institute{Princeton University, Princeton NJ 08544, USA \and
+Springer Heidelberg, Tiergartenstr. 17, 69121 Heidelberg, Germany
+\email{lncs@springer.com}\\
+\url{http://www.springer.com/gp/computer-science/lncs} \and
+ABC Institute, Rupert-Karls-University Heidelberg, Heidelberg, Germany\\
+\email{\{abc,lncs\}@uni-heidelberg.de}}
+%
+\maketitle              % typeset the header of the contribution
+%
+\begin{abstract}
+The abstract should briefly summarize the contents of the paper in
+150--250 words.
 
-\preprint{XXXXX (preprint)}
-
-%\title[Evolution of interaction networks]{On the evolution of interaction networks: primitive typology of vertex, prominence of measures and activity statistics}% Force line breaks with \\
-%\title[Evolution of interaction networks]{On the evolution of interaction networks: a primitive typology of vertex}% Force line breaks with \\
-\title[Versinus visualization of graphs]{Versinus: a visualization method for graphs in evolution}% Force line breaks with \\
-
-\author{Renato Fabbri}%
- \homepage{http://ifsc.usp.br/~fabbri/}
- \email{fabbri@usp.br}
-  \affiliation{ 
-S\~ao Carlos Institute of Physics, University of S\~ao Paulo (IFSC/USP)%\\This line break forced with \textbackslash\textbackslash
-}%
-%
-%\author{Camilo Akimushkin Valencia}
-%  \homepage{http://ifsc.usp.br/}
-%  \email{akimushkin@ifsc.usp.br}
-%  \altaffiliation[Also at ]{IFSC-USP}%Lines break automatically or can be forced with \\
+\keywords{First keyword  \and Second keyword \and Another keyword.}
+\end{abstract}
 %
 %
-%\author{Vilson Vieira da Silva Junior}
-%  \homepage{http://automata.cc/}
-%  \email{vilson@void.cc}
-%  \altaffiliation[Also at ]{IFSC-USP}%Lines break automatically or can be forced with \\
 %
-%\author{Ricardo Fabbri}
-%  \homepage{http://www.lems.brown.edu/~rfabbri/}
-%  \email{rfabbri@iprj.uerj.br}
-% \altaffiliation{
-%Instituto Polit\'ecnico, Universidade Estadual do Rio de Janeiro (IPRJ)
-%}%Lines break automatically or can be forced with \\
-%
-%\author{Deborah Christina Antunes}
-%  \homepage{http://lattes.cnpq.br/1065956470701739}
-%  \email{deborahantunes@gmail.com}
-%  \altaffiliation{
-%Curso de Psicologia, Universidade Federal do Cer\'a (UFC)
-%}%Lines break automatically or can be forced with \\
-%
-%\author{Marilia Mello Pisani}
-%  \homepage{http://lattes.cnpq.br/6738980149860322}
-%  \email{marilia.m.pisani@gmail.com}
-% \altaffiliation{
-%Centro de Ciências Naturais e Humanas, Universidade Federal do ABC (CCNH/UFABC)
-%}%Lines break automatically or can be forced with \\
-%
-%
-%\author{Luciano da Fontoura Costa}
-%  \homepage{http://cyvision.ifsc.usp.br/~luciano/}
-%  \email{ldfcosta@gmail.com}
-%  \altaffiliation[Also at ]{IFSC-USP}%Lines break automatically or can be forced with \\
-%
-%\author{Osvaldo N. Oliveira Jr.}
-%  \homepage{www.polimeros.ifsc.usp.br/professors/professor.php?id=4}
-%  \email{chu@ifsc.usp.br}
-% \altaffiliation[Also at ]{IFSC-USP}%Lines break automatically or can be forced with \\
-%
+\section{First Section}
+\subsection{A Subsection Sample}
+Please note that the first paragraph of a section or subsection is
+not indented. The first paragraph that follows a table, figure,
+equation etc. does not need an indent, either.
 
-\date{\today}% It is always \today, today,
-             %  but any date may be explicitly specified
+Subsequent paragraphs, however, are indented.
 
-\begin{abstract}
-This article presents a novel visualization approach for dynamic graphs, the versinus method,
-specially useful for real world networks exhibiting free-scale properties.
-With a simple and fixed layout, and a small set of visual markups, the method has been useful for understanding network dynamics.
-Local community often suggests that it be reported, which motivated this article.
-Online resources deliver videos and computer scripts for rendering new animations.
-This article has a concise description of the method.
-\end{abstract}
+\subsubsection{Sample Heading (Third Level)} Only two levels of
+headings should be numbered. Lower level headings remain unnumbered;
+they are formatted as run-in headings.
 
-\pacs{01.50.Fr,89.75.Fb,05.65.+b,89.65.-s}% PACS, the Physics and Astronomy
-\keywords{complex networks, visualization, social network analysis, pattern recognition, statistics}%Use showkeys class option if keyword
-\maketitle
+\paragraph{Sample Heading (Fourth Level)}
+The contribution should contain no more than four levels of
+headings. Table~\ref{tab1} gives a summary of all heading levels.
 
-%\begin{quotation}
-%% `The reason for the persistent plausibility of the typological approach, however, is not a static biological one, but just the opposite: dynamic and social.
-The fact that human society has been up to now divided into classes affects more than the external relations of men.
-The marks of social repression are left within the individual soul.'
-%% 
-%`The conception of personality structure is the
-%best safeguard against the
-%inclination to attribute persistent trends in the
-%individual to something
-%"innate" or "basic" or "racial" within him. The
-%Nazi allegation that natural, biological traits decide the total being of a person
-%would not have been such
-%a successful political device
-%had it not been possible to point to numerous
-%instances of relative fixity in human behavior and to
-%challenge those who
-%thought to explain them on any basis other than a biological one.'
-%
-%\emph{- Adorno et al, 1969, p. 747}
-%\end{quotation}
+\begin{table}
+\caption{Table captions should be placed above the
+tables.}\label{tab1}
+\begin{tabular}{|l|l|l|}
+\hline
+Heading level &  Example & Font size and style\\
+\hline
+Title (centered) &  {\Large\bfseries Lecture Notes} & 14 point, bold\\
+1st-level heading &  {\large\bfseries 1 Introduction} & 12 point, bold\\
+2nd-level heading & {\bfseries 2.1 Printing Area} & 10 point, bold\\
+3rd-level heading & {\bfseries Run-in Heading in Bold.} Text follows & 10 point, bold\\
+4th-level heading & {\itshape Lowest Level Heading.} Text follows & 10 point, italic\\
+\hline
+\end{tabular}
+\end{table}
 
-\section{\label{sec:into}Introduction}
-Versinus is a visualization method for dynamic graphs based on experimental observations.
-This method receives this dedicated article by recurrence of the suggestion, by fellow researchers,
-to write it.
+\noindent Displayed equations are centered and set on a separate
+line.
+\begin{equation}
+x + y = z
+\end{equation}
+Please try to avoid rasterized images for line-art diagrams and
+schemas. Whenever possible, use vector graphics instead (see
+Fig.~\ref{fig1}).
 
-In visualizing a network, the method consists of creating an animation,
-of a fixed-size message sliding window (e.g. 400 messages), and 
-partitioning the network in two fixed-layout segments:
-a sinusoid for the most connected vertexes
-and a straight line for the less connected.
-A vertex holds the same position throughout the animation. Also,
-visual indication of properties - such as color, height and width,
-and rank of vertex with degree criteria - plays a central role.
-This work differs from the few works on the visualization of dynamic
-graphs because it is a simple report on a method that has been useful,
-was developed by practical needs, is the result of experimentations,
-and has a number of criteria that guided its development~\cite{Viz1,Viz2,Viz3}.
-
-Next section further exposes the context in which versinus is being useful,
-and the characteristics of the systems in which it was applied.
-Section~\ref{vmethod} describes the method itself. Section~\ref{res} is dedicated
-to most important results and discussions. Conclusions and further work
-are in Section~\ref{conc}. Appendix exposes online resources.
-
-\section{Context and materials}\label{cmaterials}
-In 2013, while doing complex networks research, within a physics department,
-to visualize networks undergoing its evolution along time was a necessity.
-This posed some challenges (e.g. proper parametrization) to which two
-visualization methods were developed. The first is a web gadget, accessible via
-usual web browsers, that could render image galleries with
-measures and the node-edge structure~\cite{galGMANE, appGMANE}.
-The second method 
-is based on rendering animations with specific characteristics.
-This second method has been named \emph{versinus} for the purpose of this article.
-
-\subsection{GMANE email lists interaction networks}\label{intNet}
-The networks undergoing analysis were email interaction networks.
-Messages were obtained from the GMANE public database~\cite{GMANE,GMANE2,GMANEwikipedia}.
-For this goal, just three metadata fields of each email message is needed: the sender, the ID of the message and the ID of the message it is a response to, if any.
-Each sender is a vertex in the network.
-If a message is a response, it yields an edge from the author of the original message to current message author.
-If the edge is already there, the weight is increased by one (an edge is created with weight one).
-By inspecting all messages in a set, the interaction network given by those messages can be attained.
-This interaction network is understood as an ``information network'', with edges along the information flow, as the response is an evidence that the responder absorbed the information given by the original sender.
-If the edges are inverted, it is usual to call it a ``status network'', with the responder "giving status" to the original sender, as he considered the message worthy enough to respond it.
-
-\subsection{Sliding message window dynamic interaction networks}\label{sec:sliding}
-Consider $\Delta$ a fixed number of consecutive messages (e.g.
-$\Delta=400$).
-Consider sets $s_{i}^{i+\Delta}$ of $\Delta$ consecutive email messages, as they are registered along time.
-A Sequence $S^{\Delta,M}$ of such sets, with the first message positioned in each the $M$ considered messages (e.g.
-$M=20000$), can be written as:
-
-\begin{equation}
-S^{\Delta,M}=\{s_i^{i+\Delta}\}_{i=0}^{M-\Delta}
-\end{equation}
-
-Each set $s_i$ yields an interaction network, as described in Section~\ref{intNet}.
-Each of these sets exhibits stable properties, while each participant exhibits a wide variation of characteristics along $S$.
-To understand the mechanisms of this compatibility (unstable vertexes, stable network) led to experimenting a series of layouts and visualization techniques, from which versinus emerged.
-
-\section{The versinus visualization method}\label{vmethod}
-Versinus is a simple visualization method.
-Network vertexes are roughly split to usual $80\%$ periphery, $15\%$ intermediary and $5\%$ of hubs.
-Hubs are laid on the first half of a sinusoid.
-Intermediary on the second.
-Peripheral sector is laid on a straight line.
-This configuration can be improved in various forms, to which Section~\ref{sec:ref} is dedicated.
-Figure~\ref{fig:versinus} has an image of such a layout.
-The position of vertex, fixed, is defined by the overall structure, i.e.
-with respect to all $M$ messages (see Section~\ref{sec:sliding}).
-Numbers with individual measures for each vertex blink periodically.
-
-\begin{figure}[h!]
-    \begin{center}
-        \includegraphics[scale=.25]{figs/versinus_}
-        \caption{The versinus visualization method in use.
-5\% of The most connected vertexes (hubs) are on the left period of the sinusoid.
-15\% of the most connected remaining vertexes are on the right period.
-80\% of the least connected vertexes are on the straight line, above the sinusoidal shape.
-White dots with numbers keep track of node position in the overall degree ordering.
-Measures blink periodically on the vertex it is related to.}
-        \label{fig:versinus}
-    \end{center}
+\begin{figure}
+\includegraphics[width=\textwidth]{versinus_}
+\caption{A figure caption is always placed below the illustration.
+Please note that short captions are centered, while long ones are
+justified by the macro package automatically.} \label{fig1}
 \end{figure}
 
-\section{Results and discussion}\label{res}
-Results from versinus is divided in two groups:
-observations on features that made it useful for the task,
-and the network structure properties it made possible to grasp.
+\begin{theorem}
+This is a sample theorem. The run-in heading is set in bold, while
+the following text appears in italics. Definitions, lemmas,
+propositions, and corollaries are styled the same way.
+\end{theorem}
+%
+% the environments 'definition', 'lemma', 'proposition', 'corollary',
+% 'remark', and 'example' are defined in the LLNCS documentclass as well.
+%
+\begin{proof}
+Proofs, examples, and remarks have the initial word in italics,
+while the following text appears in normal font.
+\end{proof}
+For citations of references, we prefer the use of square brackets
+and consecutive numbers. Citations using labels or the author/year
+convention are also acceptable. The following bibliography provides
+a sample reference list with entries for journal
+articles~\cite{ref_article1}, an LNCS chapter~\cite{ref_lncs1}, a
+book~\cite{ref_book1}, proceedings without editors~\cite{ref_proc1},
+and a homepage~\cite{ref_url1}. Multiple citations are grouped
+\cite{ref_article1,ref_lncs1,ref_book1},
+\cite{ref_article1,ref_book1,ref_proc1,ref_url1}.
+%
+% ---- Bibliography ----
+%
+% BibTeX users should specify bibliography style 'splncs04'.
+% References will then be sorted and formatted in the correct style.
+%
+% \bibliographystyle{splncs04}
+% \bibliography{mybibliography}
+%
+\begin{thebibliography}{8}
+\bibitem{ref_article1}
+Author, F.: Article title. Journal \textbf{2}(5), 99--110 (2016)
 
-\subsection{Understanding of useful visualization features for dynamic networks}
+\bibitem{ref_lncs1}
+Author, F., Author, S.: Title of a proceedings paper. In: Editor,
+F., Editor, S. (eds.) CONFERENCE 2016, LNCS, vol. 9999, pp. 1--13.
+Springer, Heidelberg (2016). \doi{10.10007/1234567890}
 
-On the numerous insights related to versinus, a few of them seem more fundamental, or plain useful. This is an attempt to present them in an importance-first order:
-\begin{enumerate}
-    \item Vertexes need to remain static. Even if they move smoothly, one notices solely transient artifacts if nodes are moving around.
-    \item Very connected sectors (hubs and intermediary) need to be laid in a curve, otherwise the edges enclose each other and reasoning about the network becomes harsh.
-    \item Height and width of a vertex are very informative, specially if measures mapped to them have a strong relation, such as out-degree (mapped to height in versinus) and in-degree (mapped to width).
-    \item Coloring nodes was also informative, although less than height and weight, as these are related to measure differences more directly.
-    \item An ordering of nodes, related to their fixed position, is very useful.
-Among all tests, ordering of vertexes by degree was considered the most informative order, which led to the hub, intermediary and peripheral sectioning.
-As node position in layout is fixed, its ordering is done with respect to the overall structure, i.e.
-considering all $M$ messages.
-    \item  Numbering these positions with respect to the order of the vertexes in the larger structure (all $M$ messages) is useful in understanding how much a vertex holds his position in his scale and on a larger window size.
-\end{enumerate}
+\bibitem{ref_book1}
+Author, F., Author, S., Author, T.: Book title. 2nd edn. Publisher,
+Location (1999)
 
-Many other insights were given by versinus, such as possible visualization tools, other kind of convenient layouts and glyph elaborations. These received dedicated attention in Section~\ref{sec:ref}.
-Even so, those insights, numbered above, were incorporated to versinus as the result of tests which presented clear benefits within the best settings achieved.
+\bibitem{ref_proc1}
+Author, A.-B.: Contribution title. In: 9th International Proceedings
+on Proceedings, pp. 1--2. Publisher, Location (2010)
 
-\subsection{Understanding of network properties through versinus}
-A number of hypothesis were drawn about networks, for which versinus
-was designed. Another number of hypothesis were driven from versinus use itself.
-
-\subsubsection{Hypothesis evidenced by versinus}
-As suggested by Palla, Barab\'asi and Vicsek~\cite{barabasiEvo}, stability of participant activity in social networks is more incident in smaller networks.
-In accordance with this result, all hubs have intermittent activity in the settings analyzed, except for the email list with the smallest number of participants (the Metareciclagem email list).
-The intermitence of hubs itself was one of the top hypothesis which motivated versinus development.
-The stability of the structure~\cite{evoSN} required further verification of the instability of each participant activity.
-
-\subsubsection{Hypothesis drawn from versinus use}
-There seem to be modes of operation of the network.
-As an example, the intermediary is observed to communicate mostly to hubs or to communicate substantially with the peripheral vertex.
-Another hypothesis is that in and out degree discrepancies in each vertex are important for network characterization.
-This was confirmed by a recent work by the present author~\cite{evoSN}.
-Other hypothesis, such as discrepancies in authority and connectivity of vertex, are numerous but need further research to be valuable.
-
-\section{Conclusions and further work}\label{conc}
-Versinus has been calling the attention of fellow
-researchers for its simplicity and utility. This concise
-exposition of the method is aimed at giving further
-consequence for this enthusiasm. Results are satisfactory
-in the domain of interest and applications of the method
-to visualize other domain dynamic networks should come in near future.
-
-\subsection{Refinements on versinus}\label{sec:ref}
-By far, the most proficuous facet of versinus is its ability
-to bring insights about how to enhance its layout and use.
-Besides the technological proposal of making a tool for using versinus
-in real-time, there are some ideas about the layout and visual guides itself.
-
-To further enable visualization of hubs and intermediary vertex,
-the sinusoid can have many periods 
-with a decaying frequency.
-The upper straight line can also have an oscillating trace.
-The two halves of the sinusoidal period can be moved independently.
-The waveform need not to be a sinusoid.
-One can think of many ways to make more informative glyphs.
-Also, visual and auditory signals for specific occurrences can be interesting (e.g.
-when a new vertex appears, when one vanishes, when an ordering of vertexes changes).
-Measures of each vertex can be exposed with a vertical displacement, to enable multiple measures, to avoid the necessity to blink the numbers and to keep network visualization free from occlusion.
-
-\subsection{Other strategies envisioned}\label{sec:other}
-Working with versinus has suggested other kinds of layout for vertexes, specially other geometric figures and iterative force-based methods for positioning vertex in a fixed layout.
-Matrix visualization of graphs and recent approaches~\cite{Viz1} has been foreseen as support to versinus.
-
-\appendix
-
-\section{Online videos}
-There are video playlists online for use by the research team. Authors
-choose~\cite{animacoes} as an example. Other playlists can be found by inspecting the user.
-
-\section{Using online scripts for rendering versinus animations}
-For making the animations and the analysis of the networks, Python scripts are online in a public repository~\cite{scriptsFim}.
-
-\vspace{5mm}
-\nocite{*}
-\bibliography{paper}% Produces the bibliography via BibTeX.
-
+\bibitem{ref_url1}
+LNCS Homepage, \url{http://www.springer.com/lncs}. Last accessed 4
+Oct 2017
+\end{thebibliography}
 \end{document}
-%
-% ****** End of file aipsamp.tex ******

### 2019-05-08 doc/paper.tex
caminhado com o inicio

@@
 \usepackage{graphicx}
+\usepackage[utf8]{inputenc}
 % Used for displaying a sample figure. If possible, figure files should
@@
 %
-\title{Contribution Title\thanks{Supported by organization x.}}
+\title{Versinus: an Animated Visualization Method for Evolving Networks\thanks{Supported by FAPESP}}
 %
@@
 %
-\author{First Author\inst{1}\orcidID{0000-1111-2222-3333} \and
-Second Author\inst{2,3}\orcidID{1111-2222-3333-4444} \and
-Third Author\inst{3}\orcidID{2222--3333-4444-5555}}
+\author{Renato Fabbri\inst{1}\orcidID{0000-0002-9699-629X} \and
+Maria Cristina Ferreira de Oliveira\inst{1}\orcidID{0000-0002-4729-5104}}
 %
-\authorrunning{F. Author et al.}
+\authorrunning{R. Fabbri and M. C. F. de Oliveira}
 % First names are abbreviated in the running head.
@@
 %
-\institute{Princeton University, Princeton NJ 08544, USA \and
-Springer Heidelberg, Tiergartenstr. 17, 69121 Heidelberg, Germany
-\email{lncs@springer.com}\\
-\url{http://www.springer.com/gp/computer-science/lncs} \and
-ABC Institute, Rupert-Karls-University Heidelberg, Heidelberg, Germany\\
-\email{\{abc,lncs\}@uni-heidelberg.de}}
+\institute{University of São Paulo, São Carlos SP, BR\\
+\email{renato.fabbri@gmail.com},
+\email{cristina@icmc.usp.br}\\
+\url{http://conteudo.icmc.usp.br/pessoas/cristina/}}
 %
@@
 \begin{abstract}
-The abstract should briefly summarize the contents of the paper in
-150--250 words.
+% The abstract should briefly summarize the contents of the paper in
+% 150--250 words.
+Most real world networks change over time with new nodes and links or their removal, or even with metadata modification.
+Increasing interest in modeling evolving networks has been reported and
+a number of methods have been proposed to visualize them.  
+  This article presents a novel visualization approach for (time-)evolving (or longitudinal) networks,
+  specially useful for the observation of (in)stability of basic topological characteristics.
+  The method has received a few software implementations and has been useful in research and publication, which motivated the writing of this present document.
+  Named Versinus, the visualization consists essentially in placing the most connected nodes (hubs and intermediary) along a sine curve, and the peripheral nodes nodes along a separate segment, usually a line.
+  Thus its name from  Latin \emph{versus} (line) and \emph{sinus} (sine).
+  The method has provided insights into network properties and for network visualizations, which are described here after a formal presentation of the visualization and yielded software.
 
-\keywords{First keyword  \and Second keyword \and Another keyword.}
+\keywords{Network visualization  \and Evolving networks \and Animated visualization \and Complex networks \and Data visualization.}
 \end{abstract}
@@
 %
-\section{First Section}
-\subsection{A Subsection Sample}
-Please note that the first paragraph of a section or subsection is
-not indented. The first paragraph that follows a table, figure,
-equation etc. does not need an indent, either.
+\section{Introduction}
+The visualization of evolving networks poses challenges related to data size and complexity which have merited various contributions~\cite{evo1,evo2,evo3}.
+\subsection{Related work}
+% on animated visualization
+% on vis of evo nets
+Although animation is often criticized in the data visualization canonical literature~\cite{munzner,ware}, advantages have also been reported~\cite{cog,anim} e.g.
+making visuallizations more eye catching and due to the fact that we are accustumed to moving through the world and to moving objects.
+Evolving network visualization has been tackled both through animated and static (often timeline-based) visualizations~\cite{ego,brain,visAn}, focusing on various global and local network features.
+The contribution reported in this document is singular in consisting basically of a network layout which is fit for scale-free characteristics, together with some visualization artifacts dependent of the implementation.
+The collection of procedures and conceptualizations involved in the achieving the final visualization were also considered worth documenting as were not found elsewhere in the literature, designed simple and genuine, and found of scientific relevance~\cite{stab}.
 
-Subsequent paragraphs, however, are indented.
+\subsection{Nomenclature}
 
-\subsubsection{Sample Heading (Third Level)} Only two levels of
-headings should be numbered. Lower level headings remain unnumbered;
-they are formatted as run-in headings.
-
-\paragraph{Sample Heading (Fourth Level)}
-The contribution should contain no more than four levels of
-headings. Table~\ref{tab1} gives a summary of all heading levels.
-
-\begin{table}
-\caption{Table captions should be placed above the
-tables.}\label{tab1}
-\begin{tabular}{|l|l|l|}
-\hline
-Heading level &  Example & Font size and style\\
-\hline
-Title (centered) &  {\Large\bfseries Lecture Notes} & 14 point, bold\\
-1st-level heading &  {\large\bfseries 1 Introduction} & 12 point, bold\\
-2nd-level heading & {\bfseries 2.1 Printing Area} & 10 point, bold\\
-3rd-level heading & {\bfseries Run-in Heading in Bold.} Text follows & 10 point, bold\\
-4th-level heading & {\itshape Lowest Level Heading.} Text follows & 10 point, italic\\
-\hline
-\end{tabular}
-\end{table}
-
-\noindent Displayed equations are centered and set on a separate
-line.
-\begin{equation}
-x + y = z
-\end{equation}
-Please try to avoid rasterized images for line-art diagrams and
-schemas. Whenever possible, use vector graphics instead (see
-Fig.~\ref{fig1}).
-
-\begin{figure}
-\includegraphics[width=\textwidth]{versinus_}
-\caption{A figure caption is always placed below the illustration.
-Please note that short captions are centered, while long ones are
-justified by the macro package automatically.} \label{fig1}
-\end{figure}
-
-\begin{theorem}
-This is a sample theorem. The run-in heading is set in bold, while
-the following text appears in italics. Definitions, lemmas,
-propositions, and corollaries are styled the same way.
-\end{theorem}
-%
-% the environments 'definition', 'lemma', 'proposition', 'corollary',
-% 'remark', and 'example' are defined in the LLNCS documentclass as well.
-%
-\begin{proof}
-Proofs, examples, and remarks have the initial word in italics,
-while the following text appears in normal font.
-\end{proof}
-For citations of references, we prefer the use of square brackets
-and consecutive numbers. Citations using labels or the author/year
-convention are also acceptable. The following bibliography provides
-a sample reference list with entries for journal
-articles~\cite{ref_article1}, an LNCS chapter~\cite{ref_lncs1}, a
-book~\cite{ref_book1}, proceedings without editors~\cite{ref_proc1},
-and a homepage~\cite{ref_url1}. Multiple citations are grouped
-\cite{ref_article1,ref_lncs1,ref_book1},
-\cite{ref_article1,ref_book1,ref_proc1,ref_url1}.
+\section{A description of the method}
+\section{Software implementations}
+\section{Insights gathered}
+\section{Conclusions and further work}
 %
@@
 %
-% \bibliographystyle{splncs04}
-% \bibliography{mybibliography}
+\bibliographystyle{splncs04}
+\bibliography{paperbib}
 %
-\begin{thebibliography}{8}
-\bibitem{ref_article1}
-Author, F.: Article title. Journal \textbf{2}(5), 99--110 (2016)
-
-\bibitem{ref_lncs1}
-Author, F., Author, S.: Title of a proceedings paper. In: Editor,
-F., Editor, S. (eds.) CONFERENCE 2016, LNCS, vol. 9999, pp. 1--13.
-Springer, Heidelberg (2016). \doi{10.10007/1234567890}
-
-\bibitem{ref_book1}
-Author, F., Author, S., Author, T.: Book title. 2nd edn. Publisher,
-Location (1999)
-
-\bibitem{ref_proc1}
-Author, A.-B.: Contribution title. In: 9th International Proceedings
-on Proceedings, pp. 1--2. Publisher, Location (2010)
-
-\bibitem{ref_url1}
-LNCS Homepage, \url{http://www.springer.com/lncs}. Last accessed 4
-Oct 2017
-\end{thebibliography}
+% \begin{thebibliography}{8}
+% \bibitem{ref_article1}
+% Author, F.: Article title. Journal \textbf{2}(5), 99--110 (2016)
+% 
+% \bibitem{ref_lncs1}
+% Author, F., Author, S.: Title of a proceedings paper. In: Editor,
+% F., Editor, S. (eds.) CONFERENCE 2016, LNCS, vol. 9999, pp. 1--13.
+% Springer, Heidelberg (2016). \doi{10.10007/1234567890}
+% 
+% \bibitem{ref_book1}
+% Author, F., Author, S., Author, T.: Book title. 2nd edn. Publisher,
+% Location (1999)
+% 
+% \bibitem{ref_proc1}
+% Author, A.-B.: Contribution title. In: 9th International Proceedings
+% on Proceedings, pp. 1--2. Publisher, Location (2010)
+% 
+% \bibitem{ref_url1}
+% LNCS Homepage, \url{http://www.springer.com/lncs}. Last accessed 4
+% Oct 2017
+% \end{thebibliography}
 \end{document}

### 2019-05-08 doc/paper.tex
terminada intro v0

@@
 The visualization of evolving networks poses challenges related to data size and complexity which have merited various contributions~\cite{evo1,evo2,evo3}.
+In fact, the meaningful visualization of time-evolving networks is a well-known problem in information visualization (\emph{infovis}) and dynamic network analysis (\emph{DNA}).
+In this work, we present a novel approach in DNA infovis that consists in distributing the nodes along a fixed layout that consists in three segments for three sections (or partitions) of a scale-free network, namely for the hub, intermediary and peripheral nodes~\cite{stab}.
+The development of the network is expressed by changes in node height, width, color, format, and by changes in link disposition, thickness and color.
+Also, auxiliary global and local information are be displayed in order to enable a sound analysis of the features of interest.
+The method has been called Versinus since its early usages in 2013, because the layout stabilized, after trying many figures, in a sine period and a line, which in Latin is \emph{sinus} (sine) and \emph{versus} (line).
+Versinus has received four software implementations, two of them written in Python scripts that yield animations through stop-motion.
+The other two are written in JavaScript, one of them is an exploitation for static networks, the other is an ongoing Visual Analytics interface which extends the description in this document for 3D with additional analysis techniques such as principal component analysis (PCA).
+The writing of this document was motivated by its usefulness in scientific research~\cite{stab,thesis} and the benefit of a careful mathematical and technological description of the procedures involved in applying the method to real networks.
+
+The article is organized as follows.
+Next subsections hold remarks about related work and vocabulary.
+Section~\ref{sec:des} describes the method in detail and sufficient formalism.
+Section~\ref{sec:imp} is depicts the software implementations of Versinus.
+Section~\ref{sec:ins} is dedicated to revealing further insights obtained through the use of the method.
+Finally, Section~\ref{sec:con} convene concluding and further work commentaries.
+
 \subsection{Related work}
@@
 
-\subsection{Nomenclature}
+\subsection{Nomenclature remarks}
+Evolving networks are also called e.g. logitudinal networks, time-evolving networks, dynamic networks.
+We chose the term ``evolving network'' herein because it is simple, is semantically reasonable, and has been very much in use.
+The other main issue in current vocabulary is in the phrase ``section of a network'' (e.g.
+of the hubs).
+In most cases, the sections are also a partitions (i.e.
+they are non-overlaping sets whose union is the complete superset), but it depends of the criteria for obtaining the sectors~\cite{stab}, thus the use of the more general vocable \emph{section}.
+Moreover, network science is very multidisciplinary and context-dependent, and flourished from the ill-defined area of complex systems, favoring polysemy and synonymy.
+We strove to avoid such issues because they are usually not convenient in scientific literature.
 
-\section{A description of the method}
-\section{Software implementations}
-\section{Insights gathered}
-\section{Conclusions and further work}
+\section{A description of the method}\label{sec:des}
+\section{Software implementations}\label{sec:imp}
+\section{Insights gathered}\label{sec:ins}
+\section{Conclusions and further work}\label{sec:con}
 %

### 2019-05-09 doc/paper.tex
encaminhada descricao do posicionamento

@@
 \section{A description of the method}\label{sec:des}
+Versinus is essentially obtained through placing the nodes in segments
+with respect to their overall topological features, and then representing
+instantaneous topological features though gliph and arrow attributes.
+The visualization may be divided thus in three stages, described in the following subsections.
+
+\subsection{Node placement}
+First, a network is constructed with all the activity of interest, i.e. all the transactions, such as all the email messages considered or all the links in a data file.
+The nodes are then ordered either by degree or strength (or in- or out- degree or strength), from greatest to smallest values.
+Also from such global network, the hubs, intermediary and peripheral sectors are obtained.
+Two methods have been proven useful for obtaining such sectors: 1) assuming fixed fractions of hubs (e.g.
+5\%), of intermediary (e.g.
+15\%) and peripheral nodes (e.g.
+80\%); 2) by comparison of the network against the Erdös-Rényi model, a not trivial procedure carefully described in~\cite{stab}.
+
+With such structures at hand, the nodes are placed along three segments.
+The hubs are placed along the first half of a sine period.
+The intermediary nodes are placed along the second half of the sine period.
+The peripheral nodes are placed along an indepentent line.
+Such placement is depicted in Figure~\ref{fig:glob}.
+
+% \begin{figure}
+% \includegraphics[width=.5\textwidth]{versinus_}
+% \caption{A figure caption is always placed below the illustration.
+% Please note that short captions are centered, while long ones are
+% justified by the macro package automatically.} \label{fig1}
+% \end{figure}
+
+The parameters of the placement are:
+\begin{itemize}
+  \item the horizontal margin, i.e. the distance from the left border to the first hubs, assumed the same as the distance from the right border to last intermediary node.
+  \item The amplitude of the sine segment.
+  \item The positions $(x0,y0)$ and $(x1,y1)$, i.e. the endpoints of the segment for the peripheral sector.
+\end{itemize}
+
+\noindent Avoiding clutter with the hubs motivated both 
+the positioning of the peripheral sector in the upper-right corner and
+the placement of the most active peripheral nodes next to the intermediary sector, i.e. the ordering along the peripheral line.
+
+\subsection{Drawing nodes, links and auxiliary information}
+\subsection{Observation of the evolving network}
+
 \section{Software implementations}\label{sec:imp}

### 2019-05-09 doc/paper.tex
feita secao de posicionamento dos vertices

@@
 The peripheral nodes are placed along an indepentent line.
-Such placement is depicted in Figure~\ref{fig:glob}.
-
-% \begin{figure}
-% \includegraphics[width=.5\textwidth]{versinus_}
-% \caption{A figure caption is always placed below the illustration.
-% Please note that short captions are centered, while long ones are
-% justified by the macro package automatically.} \label{fig1}
-% \end{figure}
-
-The parameters of the placement are:
+Such placement is depicted in Figure~\ref{fig:glob} with
+the basic parameters:
 \begin{itemize}
-  \item the horizontal margin, i.e. the distance from the left border to the first hubs, assumed the same as the distance from the right border to last intermediary node.
-  \item The amplitude of the sine segment.
-  \item The positions $(x0,y0)$ and $(x1,y1)$, i.e. the endpoints of the segment for the peripheral sector.
+  \item the amplitude $\alpha$ of the sine oscillation.
+  \item The displacement $\Delta$ of the sine in the vertical axis.
+  \item the horizontal margin $\mu$, i.e. the distance from the left border to the first hubs, assumed the same as the distance from the right border to last intermediary node.
+  \item The positions $(x_0,y_0)$ and $(x_1,y_1)$, i.e. the endpoints of the segment for the peripheral sector.
 \end{itemize}
 
-\noindent Avoiding clutter with the hubs motivated both 
+\begin{figure}[!h]\centering
+\includegraphics[width=.7\textwidth]{nodePositioning}
+  \caption{The underlying layout of Versinus.
+Most connected nodes are positioned from left to right on the sine segment (for the hubs and intermediary sectors), and from right to left on the upper line (for the periphery sector).
+The parameters are: amplitude $\alpha$ of the oscillation; vertical displacement $\Delta$ of the oscillation; horizontal margin $\mu$; and line endpoints $(x_0, y_0)$ and $(x_1, y_1)$.}\label{fig1}
+\end{figure}
+
+Some crafty notes: fine-tunning of the size of each sector (number of nodes) may come handy due to the density of nodes in each segment and due to their activity.
+Also, the node positioned at the middle of the sine period may be a hub or an intermediary node, and this choice often impacts final visualization clarity more than predicted by intuition.
+Finally, 
+avoiding clutter with the hubs motivated both 
 the positioning of the peripheral sector in the upper-right corner and

### 2019-05-09 doc/mkFig.py

@@ -1,0 +1,47 @@
+import pylab as p, numpy as n
+
+fig=p.figure(figsize=(6.,4.))
+p.xlim(0,1)
+p.ylim(0,1)
+
+xmargin = 0.1
+amp = 0.7
+offset = 0.1
+x = n.linspace(0+xmargin, 0.5, 1000)
+y = 0.5 + 0.5 * amp * n.sin(n.linspace(0, n.pi, 1000)) - offset
+p.plot(x, y, label='hubs')
+
+p.annotate('', (1-xmargin, 0.2), (1, 0.2), ha="center", va="center", arrowprops=dict(arrowstyle="|-|"))
+p.annotate(r'$\mu$', (0.94, 0.22), fontsize=14)
+
+# p.arrow(0.1, 0.7, 0.1, 0.2, arrowprops=dict(arrowstyle="|-|"))
+p.annotate('', (0.05, y.max()), (0.05, y.min()), ha="center", va="center", arrowprops=dict(arrowstyle="|-|"))
+p.annotate(r'$\alpha$', (0.07, y.max()*0.75), fontsize=14)
+
+p.annotate('', (0.2, 0.5), (0.2, y.min()), ha="center", va="center", arrowprops=dict(arrowstyle="|-|"))
+p.annotate(r'$\Delta$', (0.22, 0.43), fontsize=14)
+
+x = n.linspace(0.5, 1-xmargin, 1000)
+y = 0.5 + 0.5 * amp * n.sin(n.linspace(n.pi, 2*n.pi, 1000)) - offset
+p.plot(x, y, label='intermediary')
+
+x0, y0 = 0.8, 0.6
+x1, y1 = 0.4, 0.9
+
+xx = n.linspace(x0, x1, 100)
+yy = n.linspace(y0, y1, 100)
+ax = p.plot(xx, yy, label='periphery')
+
+p.annotate(r'$(x_0, y_0)$', (x0, y0+.02), fontsize=14)
+p.annotate(r'$(x_1, y_1)$', (x1-0.02, y1+0.02), fontsize=14)
+
+p.xticks([0, 0.5, 1])
+p.yticks([0, 0.5, 1])
+p.legend()
+
+p.title("Placement of nodes in Versinus", fontsize=16)
+
+p.subplots_adjust(left=0.06, bottom=0.06, right=0.97, top=0.93)
+
+p.savefig("./nodePositioning.png")
+# p.show()

### 2019-05-09 doc/paper.tex
feito texto principal da secao de desenhar nodes de links

@@
 \usepackage[utf8]{inputenc}
+\usepackage{mathtools}
+\DeclarePairedDelimiter\floor{\lfloor}{\rfloor}
 % Used for displaying a sample figure. If possible, figure files should
@@
 
-Some crafty notes: fine-tunning of the size of each sector (number of nodes) may come handy due to the density of nodes in each segment and due to their activity.
+\noindent Some crafty notes: fine-tunning of the size of each sector (number of nodes) may come handy due to the density of nodes in each segment and due to their activity.
 Also, the node positioned at the middle of the sine period may be a hub or an intermediary node, and this choice often impacts final visualization clarity more than predicted by intuition.
@@
 \subsection{Drawing nodes, links and auxiliary information}
+For visualizing the evolving network, further concepts and quantities need to be defined.
+A \emph{snapshot} $S$ consists of a number $\sigma$ time-contiguous transactions, e.g. of 100 email messages, starting at an arbitrary message.
+The separation $\eta$ between each snapshot is the number of transactions skipped from one snapshot to the next.
+In summary, the evolving network is visualized through a sliding window of $\sigma$ transactions, updated each $\eta$ transactions.
+To achieve the final animation, one need also to define a transaction rate $\omega$.
+Thus, these parameters are specified:
+\begin{itemize}
+  \item the window $\sigma$ captured at each snapshot, expressed in transactions.
+  \item The step size $\eta$ between each snapshot, expressed in transactions.
+  \item The transaction rate $\omega$, expressed in transactions per second.
+\end{itemize}
+
+\noindent The sequence of snapshots may be expressed as $\{S_i^{\sigma, \eta}\}_0^{L}$, where $L$ is the smallest integer such that $L\times \eta + \sigma > T$ and $T$ is the total number of transactions.
+Notice that the last snapshot will not necessarily have $\sigma$ transactions, but may have less than $\sigma$.
+Also, the rate $\omega$ is only applied when rendering the animation.
+E.g.
+if each snapshot $S_i$ yields an image, the stop motion animation will use $\omega/\eta$ images per second to achieve the desired rate of $\omega$ transactions per second.
+
+Finally, auxiliary information may be given by a persistent legend that displays essential information such as $\sigma$, $\eta$, $\omega$ and $T$, and keeps track of current snapshot along e.g.
+the number of nodes and edges of the snapshot.
+Also, node-specific information may be displayed by blinking the information onver the nodes periodically or displaying them with a vertical displacement.
+Figure~\ref{fig:final} examplifies two flavours of Versinus settings.
+
 \subsection{Observation of the evolving network}

### 2019-05-09 doc/paper.tex
terminada descricao do metodo v0

@@
 \usepackage[utf8]{inputenc}
-\usepackage{mathtools}
-\DeclarePairedDelimiter\floor{\lfloor}{\rfloor}
+\usepackage{subfig}
 % Used for displaying a sample figure. If possible, figure files should
@@
 
+In each snapshot, the entailed network is analysed in terms of the sectors (hubs, intermediary and periphery), in and out degree (or strength if proffered) and clustering coefficient.
+Then the visual mapping of information is performed in the following way:
+\begin{itemize}
+  \item nodes not incident in the snapshot are not shown. Nodes incident in the snapshot is given a minimum width and height.
+  \item Node sector in the snapshot is mapped to node shape (e.g. hexagon, circle and diamond for hub, intermediary and peripheral node).
+  \item Node width is incremented proportinally to its in-degree (or out-strength if preferred). Node height is incremented proportionally to its out-degree (or out-strength if preffered).
+  \item Node color is related to clustering coefficient, e.g. zero is mapped to white, one is pure red, and intermediary values are scaled accordingly.
+  \item Link weight is mapped to link width or a color scale.
+\end{itemize}
+
 Finally, auxiliary information may be given by a persistent legend that displays essential information such as $\sigma$, $\eta$, $\omega$ and $T$, and keeps track of current snapshot along e.g.
@@
 
+\begin{figure}[!h]\centering
+  \subfloat[Versinus flavour A]{{\includegraphics[width=.46\textwidth]{im31} }}%
+    \qquad
+    \subfloat[Versinus flavour B]{{\includegraphics[width=.46\textwidth]{versinus_} }}%
+    \caption{Two frames of animations achieved using the Versinus method.
+In (a), a cleaner version of Versinus is used to display a snapshot in which the only hub is the overall hub, the second overall hub is a peripheral node in the snapshot, and nodes most nodes have high clustering coefficient.
+The legend indicates that the total number of transactions is $T=3998$, the window size $\sigma=40$, the step size is $\eta=40$, the transaction rate is $\omega=4$, current snapshot starts at transaction 1240, and the snapshot has 13 nodes and 15 links.
+In (b) the individual degrees in the snapshot is blinking over the nodes, and the nodes are counted in auxiliary lines along the bottom and the near the periphery line.
+Current image/snapshot is 911 and the snapshot has 116 nodes and 262 links.}%
+    \label{fig:final}%
+\end{figure}
+
 \subsection{Observation of the evolving network}
+Within the settings described, one may notice a number of network characteristics: do nodes alter their sectors often?
+Is the overall intermediary sector in greater activity with the hubs or periphery?
+Do higher clustering really occur most often in the less connected nodes?
+Are there very unbalanced nodes with respect to in and out degree (or strength)?
+Are the number of nodes and edges stable along the network evolution?
+In fact, Versinus was partly designed to inspect network attributes due to the remarkable stablity of their overall characteristics reported in~\cite{stab}.
+In this context, it is natural to ask if such stability is also noticeable in individual nodes, and, 
+as suggested by~\cite{barabasiEvo}, all hubs were found to have intermittent activity in social networks, except for the smallest networks analysed.
 

### 2019-05-09 doc/paper.tex
terminada secao sobre as implementacoes em fostware

@@
 Most connected nodes are positioned from left to right on the sine segment (for the hubs and intermediary sectors), and from right to left on the upper line (for the periphery sector).
-The parameters are: amplitude $\alpha$ of the oscillation; vertical displacement $\Delta$ of the oscillation; horizontal margin $\mu$; and line endpoints $(x_0, y_0)$ and $(x_1, y_1)$.}\label{fig1}
+The parameters are: amplitude $\alpha$ of the oscillation; vertical displacement $\Delta$ of the oscillation; horizontal margin $\mu$; and line endpoints $(x_0, y_0)$ and $(x_1, y_1)$.
+   Further information is given in Section~\ref{sec:des}.
+  }\label{fig1}
 \end{figure}
@@
 \begin{figure}[!h]\centering
-  \subfloat[Versinus flavour A]{{\includegraphics[width=.46\textwidth]{im31} }}%
+    \subfloat[Versinus flavour A]{{\includegraphics[width=.46\textwidth, height=4.3cm]{versinus_} }}%
     \qquad
-    \subfloat[Versinus flavour B]{{\includegraphics[width=.46\textwidth]{versinus_} }}%
+  \subfloat[Versinus flavour B]{{\includegraphics[width=.46\textwidth, height=4.3cm]{im31} }}%
     \caption{Two frames of animations achieved using the Versinus method.
-In (a), a cleaner version of Versinus is used to display a snapshot in which the only hub is the overall hub, the second overall hub is a peripheral node in the snapshot, and nodes most nodes have high clustering coefficient.
+In (a) the individual degrees in the snapshot is blinking over the nodes, and the nodes are counted in auxiliary lines along the bottom and the near the periphery line.
+Current image/snapshot is 911 and the snapshot has 116 nodes and 262 links.
+   In (b), a cleaner version of Versinus is used to display a snapshot in which the only hub is the overall hub, the second overall hub is a peripheral node in the snapshot, and nodes most nodes have high clustering coefficient.
 The legend indicates that the total number of transactions is $T=3998$, the window size $\sigma=40$, the step size is $\eta=40$, the transaction rate is $\omega=4$, current snapshot starts at transaction 1240, and the snapshot has 13 nodes and 15 links.
-In (b) the individual degrees in the snapshot is blinking over the nodes, and the nodes are counted in auxiliary lines along the bottom and the near the periphery line.
-Current image/snapshot is 911 and the snapshot has 116 nodes and 262 links.}%
+   Further information is given in Section~\ref{sec:des}.
+   }%
     \label{fig:final}%
@@
 \section{Software implementations}\label{sec:imp}
+There are four software implementations of Versinus.
+Two of them are written in Python: one is a script that renders animations as in the Figure~\ref{fig:final} (a), with more cues of local structure.
+The other is a library that renders animations as in the Figure~\ref{fig:final} (a), simpler and with more cues of the overall structure.
+The other two implementations are in JavaScript and uses WebGL.
+One is for static (non-evolving) networks, implemented as part of the ccNetViz library, that aim at being as computationally inexpensive as possible.
+The other is implemented in a 3D Visual Analytics utility which is linked to more analytical techniques, such as principal component analysis, overview first - focus on demand, simultaneous highlighting of corresponding structures across images as required by user.
+These last two implementations are illustrated in Figure~\ref{fig:extra}.
+
+\begin{figure}[!h]\centering
+    \subfloat[Versinus in JavaScript A]{{\includegraphics[width=.46\textwidth, height=2.5cm]{ccNetViz} }}%
+    \qquad
+  \subfloat[Versinus in JavaScript B]{{\includegraphics[width=.46\textwidth]{visAn} }}%
+    \caption{The two implementations of Versinus in JavaScript using WebGL.
+In (a) is an image of the implementation in ccNetViz, a library with the purpose of being computationally inexpensive.
+This implementation is only for static networks.
+In (b) is an image of the implementation in a Visual Analytics utility being developed.
+This latter software render the animations in real time, with multiple controls for inspecting the network struture and controlling the animation, in conjuntion with other analysis techniques.
+Further information is given in Section~\ref{sec:imp}.}
+    \label{fig:extra}
+\end{figure}
+
 \section{Insights gathered}\label{sec:ins}

### 2019-05-09 doc/paper.tex
feita secao v0 sobre insights

@@
 There are four software implementations of Versinus.
+For oranization and brevity, these are all linked in one repository~\cite{verRepo} together with this document.
 Two of them are written in Python: one is a script that renders animations as in the Figure~\ref{fig:final} (a), with more cues of local structure.
@@
 This latter software render the animations in real time, with multiple controls for inspecting the network struture and controlling the animation, in conjuntion with other analysis techniques.
+Also, the sine oscilates both in the vertical and depth axes, as does the periphery segment.
 Further information is given in Section~\ref{sec:imp}.}
@@
 \section{Insights gathered}\label{sec:ins}
+Valueable insights were obtained by the design and use of Versinus.
+For example, the visualization of network structures are favoured by the positoning of vertices in a oscillatory pattern.
+In fact, this principle was implemented in some of the layouts available within ccNetViz, and it may be useful in other well-known layouts, such as the Hive Plot~\cite{hive}.
+Another important insight is the pertinece of fixed positions for the observation of evolving networks.
+Not only this has the potential to help ammeliorate the hairball effect~\cite{hairball}, but it is most often very difficult to keep track of individual nodes and structures when the nodes are allowed to move in the images, as in a force-based layout.
+Furthermore, reasonable parametrization (e.g.
+through $\alpha$, $\Delta$, $\mu$, $(x_0, y_0)$, $(x_1, y_1)$, $\sigma$, $\eta$ and $\omega$) are very dependent of the network and intended tasks, and calls for interactive interfaces and automated indication of initial values.
+
 \section{Conclusions and further work}\label{sec:con}
+
 %

### 2019-05-09 doc/paper.tex
finalizado v0 do artigo

@@
 
+\subsection{Network sonification}
+The implementation in Figure~\ref{fig:extra} (b) includes a sonification routine for the evolving network.
+The results were of very modest effectiveness for the user in the sense that it did not enable one to grasp network features.
+Nevertheless, it made the (audio)visualization more interesting, and it is the first time the authors appreciated a sonification of an evolving network (maybe even of any network), and is thus probably a unique multimedia representation of networks.
+Accordingly, we believe valid a description of the sonification as it was left implemented: the lower noise has central frequency proportional to first hub connectivity, at extreme left on the layout, on the start of the cosine.
+Changes each two tempos.
+The higher noise has central frequency proportional to the second hub, and also changes every two tempos.
+The third hub connectivity is proportional to the lower note, which is incident once a beat.
+The fourth hub connectivity is proportional to the higher note, one note occurring twice every half-beat.
+
 \section{Insights gathered}\label{sec:ins}
@@
 In fact, this principle was implemented in some of the layouts available within ccNetViz, and it may be useful in other well-known layouts, such as the Hive Plot~\cite{hive}.
+The use of only one period for the sine is arbitrary and was established because it entailed good visualizations for the networks and tasks at hand during its desing, but is allowed to vary in more recent implementations.
 Another important insight is the pertinece of fixed positions for the observation of evolving networks.
 Not only this has the potential to help ammeliorate the hairball effect~\cite{hairball}, but it is most often very difficult to keep track of individual nodes and structures when the nodes are allowed to move in the images, as in a force-based layout.
+The positioning of nodes with respect to their topological features (nodes are ordered by connectivity, segments are related to network sectors) revealed very useful.
+Visual cues such as node width, height, shape and color was found very informative for individual nodes but resulted in a cognitive overload for the observation of global features, which motivates simplified versions.
+The sonification of the evolving network added to the exotic flavour of Versinus and made it more attracive.
+It seems that more informative sonifications will require a dedicated work, training of already musically competent users, and very modest goals.
 Furthermore, reasonable parametrization (e.g.
@@
 \section{Conclusions and further work}\label{sec:con}
+Within this document, we achieved a first sound description of the Versinus method, with its formal elements, use context and software implementations.
+The validity of Versinus is implied both by its use in published research scenarios and the insights it yielded on individual networks and network layouts in general.
+Further developments using the method are already being made, with focus on interactivity and Web technologies, such as WebGL and the use of well-polished JavaScript libraries.
+
+Next steps include
+automated tunning of Versinus parametrization, 
+implementation of cues for relating language incident in the networks to their topology, and
+further visually representing topological features (e.g. the fraction of interaction of the intermediary sector with the hubs and with the periphery).
+The experimentation of Versinus features within other layouts is a potentially
+valuable contribution, e.g. wiggling the segments of a Hive Plot~\cite{hive}.
+Most importantly, we aim at completing the Visual Analytics utility,
+enabling the use of Versinus in an interactive environment, for enhanced exploratory analysis.
+Finally, we plan to further analysing datasets of evolving networks~\cite{stab}.
 

### 2019-05-10 doc/paper.tex
escrito sobre os videos

@@
 
+\subsection{Animations produced and available}\label{sec:vid}
+Many animations were obtained through the routines described.
+These encompassed different networks and settings ($\alpha$, $\sigma$, $\eta$,
+etc), and different flavours of Versinus.
+Some of these animations are available in six playlists of Youtube videos.
+For organization and brevity, they are also available at~\cite{verRepo}.
+
 \section{Insights gathered}\label{sec:ins}

### 2019-05-10 doc/paper.tex
mexido no readme
finalizada revisao

--- 2019-05-09 doc/paper.tex
@@
 %
-\title{Versinus: an Animated Visualization Method for Evolving Networks\thanks{Supported by FAPESP}}
+\title{Versinus: an Animated Visualization Method for Evolving Networks\thanks{Supported by FAPESP, project 2017/05838-3}}
 %
@@
 %
-\author{Renato Fabbri\inst{1}\orcidID{0000-0002-9699-629X} \and
-Maria Cristina Ferreira de Oliveira\inst{1}\orcidID{0000-0002-4729-5104}}
+\author{Renato Fabbri\orcidID{0000-0002-9699-629X} \and
+Maria Cristina Ferreira de Oliveira\orcidID{0000-0002-4729-5104}}
 %
@@
 % 150--250 words.
-Most real world networks change over time with new nodes and links or their removal, or even with metadata modification.
+Most real-world networks change over time with new nodes and links or their removal, and their representation may involve metadata variation.
 Increasing interest in modeling evolving networks has been reported and
-a number of methods have been proposed to visualize them.  
-  This article presents a novel visualization approach for (time-)evolving (or longitudinal) networks,
-  specially useful for the observation of (in)stability of basic topological characteristics.
-  The method has received a few software implementations and has been useful in research and publication, which motivated the writing of this present document.
-  Named Versinus, the visualization consists essentially in placing the most connected nodes (hubs and intermediary) along a sine curve, and the peripheral nodes nodes along a separate segment, usually a line.
-  Thus its name from  Latin \emph{versus} (line) and \emph{sinus} (sine).
+a number of procedures have been proposed to visualize them.
+  This article presents a novel visualization method for evolving (or longitudinal, or time-evolving) networks,
+  specially useful in the observation of (in)stability of basic topological characteristics.
+  The method has received a few software implementations and has been useful for research and publication, which motivated the writing of this present document.
+  Named Versinus, the visualization consists essentially in placing the most connected nodes (hubs and intermediary) along a sine curve, and the peripheral nodes along a separate segment, usually a line.
+  Thus its name, from  Latin \emph{versus} (line) and \emph{sinus} (sine).
   The method has provided insights into network properties and for network visualizations, which are described here after a formal presentation of the visualization and yielded software.
@@
 \section{Introduction}
-The visualization of evolving networks poses challenges related to data size and complexity which have merited various contributions~\cite{evo1,evo2,evo3}.
+The visualization of evolving networks poses challenges related to data size and complexity which have merited various contributions~\cite{evo1,evo3,evo2}.
 In fact, the meaningful visualization of time-evolving networks is a well-known problem in information visualization (\emph{infovis}) and dynamic network analysis (\emph{DNA}).
-In this work, we present a novel approach in DNA infovis that consists in distributing the nodes along a fixed layout that consists in three segments for three sections (or partitions) of a scale-free network, namely for the hub, intermediary and peripheral nodes~\cite{stab}.
+In this work, we present a novel approach in DNA infovis that consists in distributing the nodes along a fixed layout that consists in three segments for three sectors (or partitions) of a scale-free network, namely for the hub, intermediary and peripheral nodes~\cite{stab}.
 The development of the network is expressed by changes in node height, width, color, format, and by changes in link disposition, thickness and color.
-Also, auxiliary global and local information are be displayed in order to enable a sound analysis of the features of interest.
+Also, auxiliary global and local information are be displayed in order to enable a sound analysis with the features of interest.
 The method has been called Versinus since its early usages in 2013, because the layout stabilized, after trying many figures, in a sine period and a line, which in Latin is \emph{sinus} (sine) and \emph{versus} (line).
-Versinus has received four software implementations, two of them written in Python scripts that yield animations through stop-motion.
-The other two are written in JavaScript, one of them is an exploitation for static networks, the other is an ongoing Visual Analytics interface which extends the description in this document for 3D with additional analysis techniques such as principal component analysis (PCA).
-The writing of this document was motivated by its usefulness in scientific research~\cite{stab,thesis} and the benefit of a careful mathematical and technological description of the procedures involved in applying the method to real networks.
+Versinus has received four software implementations, two of them written in Python scripts and packages that yield animations through stop-motion.
+The other two are written in JavaScript, one of them is an exploitation for static networks, the other is an ongoing Visual Analytics interface which extends the description in this document for 3D with additional analysis techniques such as principal component analysis (PCA) and structure navigation.
+The writing of this document was motivated by the usefulness of the method in scientific research~\cite{thesis,stab} and the benefit of a careful mathematical and technological description of the procedures involved in applying the method to real networks.
 
@@
 Section~\ref{sec:des} describes the method in detail and sufficient formalism.
-Section~\ref{sec:imp} is depicts the software implementations of Versinus.
+Section~\ref{sec:imp} depicts the software implementations of Versinus and resulted animations.
 Section~\ref{sec:ins} is dedicated to revealing further insights obtained through the use of the method.
-Finally, Section~\ref{sec:con} convene concluding and further work commentaries.
+Finally, Section~\ref{sec:con} assemble concluding commentaries and further work.
 
@@
 % on vis of evo nets
-Although animation is often criticized in the data visualization canonical literature~\cite{munzner,ware}, advantages have also been reported~\cite{cog,anim} e.g.
+Although animation is often criticized in the data visualization canonical literature, advantages have also been reported~\cite{anim,munzner,cog,ware} e.g.
 making visuallizations more eye catching and due to the fact that we are accustumed to moving through the world and to moving objects.
 Evolving network visualization has been tackled both through animated and static (often timeline-based) visualizations~\cite{ego,brain,visAn}, focusing on various global and local network features.
-The contribution reported in this document is singular in consisting basically of a network layout which is fit for scale-free characteristics, together with some visualization artifacts dependent of the implementation.
-The collection of procedures and conceptualizations involved in the achieving the final visualization were also considered worth documenting as were not found elsewhere in the literature, designed simple and genuine, and found of scientific relevance~\cite{stab}.
+The contribution reported in this document is singular in consisting basically of a network layout which is fit for scale-free characteristics, together with some visualization cues dependent on the implementation, and in being the result of extensive attempts to reach a satisfactory visualization by theoretical and intuitive strategies and methodical verification of the result by appreciation of the entailed animations.
+The procedures and conceptualizations involved in the achieving the final visualization were also considered worth documenting as they were not found elsewhere in the literature, were designed simple and genuine, and found of scientific relevance~\cite{stab}.
 
@@
 We chose the term ``evolving network'' herein because it is simple, is semantically reasonable, and has been very much in use.
-The other main issue in current vocabulary is in the phrase ``section of a network'' (e.g.
+The other main issue in current vocabulary is in the phrase ``sector of a network'' (e.g.
 of the hubs).
-In most cases, the sections are also a partitions (i.e.
-they are non-overlaping sets whose union is the complete superset), but it depends of the criteria for obtaining the sectors~\cite{stab}, thus the use of the more general vocable \emph{section}.
+In most cases, the sectors are also partitions (i.e.
+they are non-overlaping sets whose union is the complete superset), but it depends of the criteria for obtaining the sectors~\cite{stab}, thus the use of the more general vocable \emph{sector}.
 Moreover, network science is very multidisciplinary and context-dependent, and flourished from the ill-defined area of complex systems, favoring polysemy and synonymy.
@@
 
-\subsection{Node placement}
+\subsection{Node placement}\label{sec:pla}
 First, a network is constructed with all the activity of interest, i.e. all the transactions, such as all the email messages considered or all the links in a data file.
@@
 15\%) and peripheral nodes (e.g.
-80\%); 2) by comparison of the network against the Erdös-Rényi model, a not trivial procedure carefully described in~\cite{stab}.
+80\%); 2) by comparison of the network against the Erdös-Rényi model, a non-trivial procedure, with relevant analytical results, carefully described in~\cite{stab}.
 
@@
 The parameters are: amplitude $\alpha$ of the oscillation; vertical displacement $\Delta$ of the oscillation; horizontal margin $\mu$; and line endpoints $(x_0, y_0)$ and $(x_1, y_1)$.
-   Further information is given in Section~\ref{sec:des}.
-  }\label{fig1}
+   Further information is given in Section~\ref{sec:pla}.
+  }\label{fig:glob}
 \end{figure}
@@
 \noindent The sequence of snapshots may be expressed as $\{S_i^{\sigma, \eta}\}_0^{L}$, where $L$ is the smallest integer such that $L\times \eta + \sigma > T$ and $T$ is the total number of transactions.
-Notice that the last snapshot will not necessarily have $\sigma$ transactions, but may have less than $\sigma$.
+Notice that the last snapshot $S_L$ will not necessarily have $\sigma$ transactions, but may have less than $\sigma$.
 Also, the rate $\omega$ is only applied when rendering the animation.
@@
   \item nodes not incident in the snapshot are not shown. Nodes incident in the snapshot is given a minimum width and height.
-  \item Node sector in the snapshot is mapped to node shape (e.g. hexagon, circle and diamond for hub, intermediary and peripheral node).
+  \item Node sector in the snapshot is mapped to node shape (e.g. hexagon, circle and diamond for hub, intermediary and peripheral nodes).
   \item Node width is incremented proportinally to its in-degree (or out-strength if preferred). Node height is incremented proportionally to its out-degree (or out-strength if preffered).
   \item Node color is related to clustering coefficient, e.g. zero is mapped to white, one is pure red, and intermediary values are scaled accordingly.
-  \item Link weight is mapped to link width or a color scale.
+  \item Link weight is mapped to link width and/or a color scale.
 \end{itemize}
@@
 the number of nodes and edges of the snapshot.
-Also, node-specific information may be displayed by blinking the information onver the nodes periodically or displaying them with a vertical displacement.
+Also, node-specific information may be displayed by blinking corresponding cues over the nodes periodically or displaying them with a vertical displacement.
 Figure~\ref{fig:final} examplifies two flavours of Versinus settings.
@@
 Current image/snapshot is 911 and the snapshot has 116 nodes and 262 links.
-   In (b), a cleaner version of Versinus is used to display a snapshot in which the only hub is the overall hub, the second overall hub is a peripheral node in the snapshot, and nodes most nodes have high clustering coefficient.
+   In (b), a cleaner version of Versinus is used to display a snapshot in which the only hub is the overall hub, the second overall hub is a peripheral node, and most nodes have high clustering coefficients.
 The legend indicates that the total number of transactions is $T=3998$, the window size $\sigma=40$, the step size is $\eta=40$, the transaction rate is $\omega=4$, current snapshot starts at transaction 1240, and the snapshot has 13 nodes and 15 links.
@@
 \subsection{Observation of the evolving network}
-Within the settings described, one may notice a number of network characteristics: do nodes alter their sectors often?
+Within the settings described, one may notice a number of network characteristics: do nodes (or a specific node) change their sectors often?
 Is the overall intermediary sector in greater activity with the hubs or periphery?
@@
 There are four software implementations of Versinus.
-For oranization and brevity, these are all linked in one repository~\cite{verRepo} together with this document.
+For organization and brevity, these are all linked in this repository:~\cite{verRepo}, together with this document.
 Two of them are written in Python: one is a script that renders animations as in the Figure~\ref{fig:final} (a), with more cues of local structure.
@@
 The other two implementations are in JavaScript and uses WebGL.
-One is for static (non-evolving) networks, implemented as part of the ccNetViz library, that aim at being as computationally inexpensive as possible.
-The other is implemented in a 3D Visual Analytics utility which is linked to more analytical techniques, such as principal component analysis, overview first - focus on demand, simultaneous highlighting of corresponding structures across images as required by user.
+One is for static (non-evolving) networks, implemented as part of the ccNetViz library, that aims at being as computationally inexpensive as possible~\cite{cell}.
+The other is implemented in an under-development 3D Visual Analytics utility in which Versinus is combined with other analytical techniques, such as principal component analysis, overview first - focus on demand, and simultaneous highlighting of corresponding structures across images as required by user.
 These last two implementations are illustrated in Figure~\ref{fig:extra}.
@@
 This implementation is only for static networks.
-In (b) is an image of the implementation in a Visual Analytics utility being developed.
-This latter software render the animations in real time, with multiple controls for inspecting the network struture and controlling the animation, in conjuntion with other analysis techniques.
+In (b) is an image of the Versinus implementation in a Visual Analytics utility being developed.
+This latter software render the animations in real time, with multiple controls for inspecting the network struture and controlling the animation, in conjunction with other analysis techniques.
 Also, the sine oscilates both in the vertical and depth axes, as does the periphery segment.
@@
 
+\subsection{Network sonification}
+The implementation illustrated in Figure~\ref{fig:final} (b) includes a sonification routine for the evolving network.
+The results were of very modest effectiveness in the sense that it does not enable the user to grasp network features with ease, specially if the user is not musically trained.
+Nevertheless, it made the (audio)visualization more interesting, and it is the first time the authors appreciated a sonification of an evolving network (maybe even of any network), and is thus probably a unique multimedia representation of networks.
+Accordingly, we believe valid a description of the sonification as it was left implemented, i.e.
+in its final version: the lower noise has central frequency proportional to first hub connectivity, at the extreme left on the layout, on the start of the sine, and it updated each two tempos.
+The higher noise has central frequency proportional to the second hub, and is also updated every two tempos.
+The third hub connectivity is proportional to the lower note, which is incident once a beat.
+The fourth hub connectivity is proportional to the higher note, one note occurring twice every half-beat.
+
+\subsection{Animations produced and available}\label{sec:vid}
+Many animations were obtained through the routines described.
+These encompassed different networks and settings ($\alpha$, $\sigma$, $\eta$,
+etc), and different flavours of Versinus.
+Some of these animations are available in six playlists of Youtube videos.
+For organization and brevity, they are available at~\cite{verRepo} together with other Versinus-related resources.
+
 \section{Insights gathered}\label{sec:ins}
 Valueable insights were obtained by the design and use of Versinus.
-For example, the visualization of network structures are favoured by the positoning of vertices in a oscillatory pattern.
+For example, the visualization of network structures are favoured by the positoning of nodes in an oscillatory pattern.
 In fact, this principle was implemented in some of the layouts available within ccNetViz, and it may be useful in other well-known layouts, such as the Hive Plot~\cite{hive}.
+The use of only one period for the sine is arbitrary and was established because it entailed good visualizations for the networks and tasks at hand during the desing of Versinus, but is allowed to vary in more recent implementations.
 Another important insight is the pertinece of fixed positions for the observation of evolving networks.
 Not only this has the potential to help ammeliorate the hairball effect~\cite{hairball}, but it is most often very difficult to keep track of individual nodes and structures when the nodes are allowed to move in the images, as in a force-based layout.
+The positioning of nodes with respect to their topological features (nodes are ordered by connectivity, segments are related to network sectors) revealed very useful for understanding network features, such as stability and connectivity dependant behaviors.
+Visual cues such as node width, height, shape and color were found informative for individual nodes but resulted in a cognitive overload for the observation of global features, which motivates simplified versions.
+The sonification of the evolving network added to the exotic flavour of Versinus and made it more attractive.
+It seems that more informative sonifications will require a dedicated work, training of already musically competent users, and very modest goals.
 Furthermore, reasonable parametrization (e.g.
@@
 \section{Conclusions and further work}\label{sec:con}
+Within this document, we achieved a first thorough description of the Versinus method, with its formal elements, use context and software implementations.
+The validity of Versinus is implied both by its use in published research scenarios and the insights it yielded on individual networks and network layouts in general.
+Further developments using the method are already being made, with focus on interactivity and Web technologies, such as WebGL and the use of well-polished JavaScript libraries.
+
+Next steps include
+automated tunning of Versinus parameters, 
+implementation of cues for relating language incident in the networks to their topology, and
+further visually representing topological features (e.g. the fraction of interaction of the intermediary sector with the hubs and with the periphery).
+The experimentation of Versinus features within other layouts is a potentially
+valuable contribution, e.g. wiggling the segments of a Hive Plot~\cite{hive}.
+Most importantly, we aim at completing the Visual Analytics utility,
+enabling the use of Versinus in an interactive environment, for enhanced exploratory analysis.
+Finally, we plan to further analyze datasets of evolving networks~\cite{stab} using Versinus.
 

### 2019-05-10 doc/paper.tex
spell check

@@
 Although animation is often criticized in the data visualization canonical literature, advantages have also been reported~\cite{anim,munzner,cog,ware} e.g.
-making visuallizations more eye catching and due to the fact that we are accustumed to moving through the world and to moving objects.
+making visualizations more eye catching and due to the fact that we are accustomed to moving through the world and to moving objects.
 Evolving network visualization has been tackled both through animated and static (often timeline-based) visualizations~\cite{ego,brain,visAn}, focusing on various global and local network features.
@@
 \subsection{Nomenclature remarks}
-Evolving networks are also called e.g. logitudinal networks, time-evolving networks, dynamic networks.
+Evolving networks are also called e.g. longitudinal networks, time-evolving networks, dynamic networks.
 We chose the term ``evolving network'' herein because it is simple, is semantically reasonable, and has been very much in use.
@@
 In most cases, the sectors are also partitions (i.e.
-they are non-overlaping sets whose union is the complete superset), but it depends of the criteria for obtaining the sectors~\cite{stab}, thus the use of the more general vocable \emph{sector}.
+they are non-overlapping sets whose union is the complete superset), but it depends of the criteria for obtaining the sectors~\cite{stab}, thus the use of the more general vocable \emph{sector}.
 Moreover, network science is very multidisciplinary and context-dependent, and flourished from the ill-defined area of complex systems, favoring polysemy and synonymy.
@@
 with respect to their overall topological features, and then representing
-instantaneous topological features though gliph and arrow attributes.
+instantaneous topological features though glyph and arrow attributes.
 The visualization may be divided thus in three stages, described in the following subsections.
@@
 The intermediary nodes are placed along the second half of the sine period.
-The peripheral nodes are placed along an indepentent line.
+The peripheral nodes are placed along an independent line.
 Such placement is depicted in Figure~\ref{fig:glob} with
@@
 
-\noindent Some crafty notes: fine-tunning of the size of each sector (number of nodes) may come handy due to the density of nodes in each segment and due to their activity.
+\noindent Some crafty notes: fine-tuning of the size of each sector (number of nodes) may come handy due to the density of nodes in each segment and due to their activity.
 Also, the node positioned at the middle of the sine period may be a hub or an intermediary node, and this choice often impacts final visualization clarity more than predicted by intuition.
@@
 
-In each snapshot, the entailed network is analysed in terms of the sectors (hubs, intermediary and periphery), in and out degree (or strength if proffered) and clustering coefficient.
+In each snapshot, the entailed network is analyzed in terms of the sectors (hubs, intermediary and periphery), in and out degree (or strength if proffered) and clustering coefficient.
 Then the visual mapping of information is performed in the following way:
@@
   \item Node sector in the snapshot is mapped to node shape (e.g. hexagon, circle and diamond for hub, intermediary and peripheral nodes).
-  \item Node width is incremented proportinally to its in-degree (or out-strength if preferred). Node height is incremented proportionally to its out-degree (or out-strength if preffered).
+  \item Node width is incremented proportionally to its in-degree (or out-strength if preferred). Node height is incremented proportionally to its out-degree (or out-strength if preferred).
   \item Node color is related to clustering coefficient, e.g. zero is mapped to white, one is pure red, and intermediary values are scaled accordingly.
@@
 Also, node-specific information may be displayed by blinking corresponding cues over the nodes periodically or displaying them with a vertical displacement.
-Figure~\ref{fig:final} examplifies two flavours of Versinus settings.
+Figure~\ref{fig:final} exemplifies two flavors of Versinus settings.
 
@@
 Are the number of nodes and edges stable along the network evolution?
-In fact, Versinus was partly designed to inspect network attributes due to the remarkable stablity of their overall characteristics reported in~\cite{stab}.
+In fact, Versinus was partly designed to inspect network attributes due to the remarkable stability of their overall characteristics reported in~\cite{stab}.
 In this context, it is natural to ask if such stability is also noticeable in individual nodes, and, 
-as suggested by~\cite{barabasiEvo}, all hubs were found to have intermittent activity in social networks, except for the smallest networks analysed.
+as suggested by~\cite{barabasiEvo}, all hubs were found to have intermittent activity in social networks, except for the smallest networks analyzed.
 
@@
 In (b) is an image of the Versinus implementation in a Visual Analytics utility being developed.
-This latter software render the animations in real time, with multiple controls for inspecting the network struture and controlling the animation, in conjunction with other analysis techniques.
-Also, the sine oscilates both in the vertical and depth axes, as does the periphery segment.
+This latter software render the animations in real time, with multiple controls for inspecting the network structure and controlling the animation, in conjunction with other analysis techniques.
+Also, the sine oscillates both in the vertical and depth axes, as does the periphery segment.
 Further information is given in Section~\ref{sec:imp}.}
@@
 These encompassed different networks and settings ($\alpha$, $\sigma$, $\eta$,
-etc), and different flavours of Versinus.
+etc), and different flavors of Versinus.
 Some of these animations are available in six playlists of Youtube videos.
@@
 \section{Insights gathered}\label{sec:ins}
-Valueable insights were obtained by the design and use of Versinus.
-For example, the visualization of network structures are favoured by the positoning of nodes in an oscillatory pattern.
+Valuable insights were obtained by the design and use of Versinus.
+For example, the visualization of network structures are favored by the positioning of nodes in an oscillatory pattern.
 In fact, this principle was implemented in some of the layouts available within ccNetViz, and it may be useful in other well-known layouts, such as the Hive Plot~\cite{hive}.
-The use of only one period for the sine is arbitrary and was established because it entailed good visualizations for the networks and tasks at hand during the desing of Versinus, but is allowed to vary in more recent implementations.
-Another important insight is the pertinece of fixed positions for the observation of evolving networks.
-Not only this has the potential to help ammeliorate the hairball effect~\cite{hairball}, but it is most often very difficult to keep track of individual nodes and structures when the nodes are allowed to move in the images, as in a force-based layout.
-The positioning of nodes with respect to their topological features (nodes are ordered by connectivity, segments are related to network sectors) revealed very useful for understanding network features, such as stability and connectivity dependant behaviors.
+The use of only one period for the sine is arbitrary and was established because it entailed good visualizations for the networks and tasks at hand during the design of Versinus, but is allowed to vary in more recent implementations.
+Another important insight is the pertinence of fixed positions for the observation of evolving networks.
+Not only this has the potential to help ameliorate the hairball effect~\cite{hairball}, but it is most often very difficult to keep track of individual nodes and structures when the nodes are allowed to move in the images, as in a force-based layout.
+The positioning of nodes with respect to their topological features (nodes are ordered by connectivity, segments are related to network sectors) revealed very useful for understanding network features, such as stability and connectivity dependent behaviors.
 Visual cues such as node width, height, shape and color were found informative for individual nodes but resulted in a cognitive overload for the observation of global features, which motivates simplified versions.
-The sonification of the evolving network added to the exotic flavour of Versinus and made it more attractive.
+The sonification of the evolving network added to the exotic flavor of Versinus and made it more attractive.
 It seems that more informative sonifications will require a dedicated work, training of already musically competent users, and very modest goals.
@@
 Next steps include
-automated tunning of Versinus parameters, 
+automated tuning of Versinus parameters, 
 implementation of cues for relating language incident in the networks to their topology, and

### 2019-06-12 doc/paper.tex
subindo versao atualizada do escrito

@@
 %
-\title{Versinus: an Animated Visualization Method for Evolving Networks\thanks{Supported by FAPESP, project 2017/05838-3}}
+\title{Versinus: an Animated Visualization Method for Dynamic Scale-Free Networks\thanks{Supported by FAPESP, project 2017/05838-3}}
 %
@@
 %
-\institute{University of São Paulo, São Carlos SP, BR\\
+\institute{University of São Paulo, São Carlos SP, Brazil\\
 \email{renato.fabbri@gmail.com},
-\email{cristina@icmc.usp.br}\\
-\url{http://conteudo.icmc.usp.br/pessoas/cristina/}}
+\email{cristina@icmc.usp.br}}
+%\url{http://conteudo.icmc.usp.br/pessoas/cristina/}}
 %
@@
 % 150--250 words.
-Most real-world networks change over time with new nodes and links or their removal, and their representation may involve metadata variation.
-Increasing interest in modeling evolving networks has been reported and
+Most real-world networks are dynamic, i.e., they change over time as nodes and links are added or removed, and their representation may also involve variations in the associated data.
+There is a growing interest in modeling dynamic (or time-evolving, or longitudinal) networks and
 a number of procedures have been proposed to visualize them.
-  This article presents a novel visualization method for evolving (or longitudinal, or time-evolving) networks,
-  specially useful in the observation of (in)stability of basic topological characteristics.
-  The method has received a few software implementations and has been useful for research and publication, which motivated the writing of this present document.
-  Named Versinus, the visualization consists essentially in placing the most connected nodes (hubs and intermediary) along a sine curve, and the peripheral nodes along a separate segment, usually a line.
+  This article presents a visualization method for dynamic scale-free networks.
+  Compliant methods were not found in other tools by the authors, which motivated this description and the development of a dedicated interface.
+  Named Versinus, the visualization is specially useful in the observation of (in)stability of basic topological characteristics, and consists essentially in placing the most connected nodes (hubs and intermediary) along a sine curve, and the peripheral nodes along a separate segment, usually a line.
   Thus its name, from  Latin \emph{versus} (line) and \emph{sinus} (sine).
-  The method has provided insights into network properties and for network visualizations, which are described here after a formal presentation of the visualization and yielded software.
+  The Versinus visualization can provide insight into network properties and is useful for observing its behavior, as described in this paper, after a formal presentation of the method and its associated software implementations.
+\keywords{Network visualization
+12
 
-\keywords{Network visualization  \and Evolving networks \and Animated visualization \and Complex networks \and Data visualization.}
+\and Dynamic networks
+\and Longitudinal networks
+\and Time-evolving networks
+\and Evolving networks
+\and Animated visualization
+\and Complex networks
+\and Data visualization.}
 \end{abstract}
@@
 \section{Introduction}
-The visualization of evolving networks poses challenges related to data size and complexity which have merited various contributions~\cite{evo1,evo3,evo2}.
-In fact, the meaningful visualization of time-evolving networks is a well-known problem in information visualization (\emph{infovis}) and dynamic network analysis (\emph{DNA}).
-In this work, we present a novel approach in DNA infovis that consists in distributing the nodes along a fixed layout that consists in three segments for three sectors (or partitions) of a scale-free network, namely for the hub, intermediary and peripheral nodes~\cite{stab}.
-The development of the network is expressed by changes in node height, width, color, format, and by changes in link disposition, thickness and color.
-Also, auxiliary global and local information are be displayed in order to enable a sound analysis with the features of interest.
-The method has been called Versinus since its early usages in 2013, because the layout stabilized, after trying many figures, in a sine period and a line, which in Latin is \emph{sinus} (sine) and \emph{versus} (line).
-Versinus has received four software implementations, two of them written in Python scripts and packages that yield animations through stop-motion.
-The other two are written in JavaScript, one of them is an exploitation for static networks, the other is an ongoing Visual Analytics interface which extends the description in this document for 3D with additional analysis techniques such as principal component analysis (PCA) and structure navigation.
-The writing of this document was motivated by the usefulness of the method in scientific research~\cite{thesis,stab} and the benefit of a careful mathematical and technological description of the procedures involved in applying the method to real networks.
+The visualization of dynamic networks poses challenges related to data size and complexity which have merited various contributions~\cite{evo1,hajij,evo3,evo2}.
+In fact, the meaningful visualization of dynamic networks is a well-known problem in Information Visualization and dynamic network analysis.
+In this work, we present an approach for dynamic network visualization that consists in distributing the nodes along a fixed layout arranged in three segments associated with the three sectors (or partitions) of a scale-free network, namely for the hub, intermediary and peripheral nodes~\cite{stab}.
+The network dynamic features are expressed by means of changes in visual properties of the glyphs used to depict nodes and links, e.g., glyph height, width, color and shape, and line (arrow) disposition, thickness and color.
+Moreover, complementary global and local information is rendered in order to enable an acoustic depiction of features of interest.
+The method has been called Versinus since its early usages in 2013, because the layout stabilized, after attempts with distinct arrangements, in a sine period and a line, which in Latin is \emph{sinus} (sine) and \emph{versus} (line).
+There are four software implementations of Versinus, two of them written in Python scripts and packages that yield animations through stop-motion.
+The other two are written in JavaScript, one focused on exploitation of static networks, the other is an experimental Visual Analytics interface, which extends the visual metaphor to the 3D space and provides additional analysis techniques such as principal component analysis (PCA) and resources for structure navigation\footnote{An instance of the interface is available at: http://rfabbri.vicg.icmc.usp.br:3000/evolution}.
+The writing of this document was motivated by
+the fact that no method was found by the authors that aims at visualizing dynamic scale-free networks through node-link diagrams.
+Furthermore, the visualization method was employed in scientific research~\cite{thesis,stab}, and thus a careful mathematical and technological description of the procedures involved in applying Versinus to real networks is appropriate.
 
 The article is organized as follows.
-Next subsections hold remarks about related work and vocabulary.
+The next subsections address related work and issues of nomenclature.
 Section~\ref{sec:des} describes the method in detail and sufficient formalism.
-Section~\ref{sec:imp} depicts the software implementations of Versinus and resulted animations.
+Section~\ref{sec:imp} outlines the software implementations of Versinus and depicts the resulting animations.
 Section~\ref{sec:ins} is dedicated to revealing further insights obtained through the use of the method.
-Finally, Section~\ref{sec:con} assemble concluding commentaries and further work.
+Finally, Section~\ref{sec:con} assembles concluding commentaries and further work.
 
@@
 % on vis of evo nets
-Although animation is often criticized in the data visualization canonical literature, advantages have also been reported~\cite{anim,munzner,cog,ware} e.g.
-making visualizations more eye catching and due to the fact that we are accustomed to moving through the world and to moving objects.
-Evolving network visualization has been tackled both through animated and static (often timeline-based) visualizations~\cite{ego,brain,visAn}, focusing on various global and local network features.
-The contribution reported in this document is singular in consisting basically of a network layout which is fit for scale-free characteristics, together with some visualization cues dependent on the implementation, and in being the result of extensive attempts to reach a satisfactory visualization by theoretical and intuitive strategies and methodical verification of the result by appreciation of the entailed animations.
-The procedures and conceptualizations involved in the achieving the final visualization were also considered worth documenting as they were not found elsewhere in the literature, were designed simple and genuine, and found of scientific relevance~\cite{stab}.
+Although animation for the purposes of data analysis is often criticized in the data visualization literature, advantages have also been reported~\cite{anim,munzner,cog,ware}, e.g.: yielding visualizations more eye catching and taking advantage of the fact that humans are used to interacting with moving objects.
+Dynamic network visualization has been tackled both through animated and static (often timeline-based) visualizations~\cite{hajij,brain,visAn,ego}, focusing on various global and local network features.
+The contribution reported in this document is singular in that it consists basically of a layout which is fit for networks with scale-free characteristics, in association with some visualization cues which are implementation dependent.
+It results from extensive attempts to reach a satisfactory visualization by means theoretical and intuitive strategies and methodical verification of the result by appreciation of the entailed animations.
+The procedures and conceptualizations involved in achieving the final visualization were also considered worth documenting, as they were not found elsewhere in the literature, and by virtue of the simple and genuine design.
+Its scientific relevance on a practical application has been demonstrated before~\cite{stab}.
 
-\subsection{Nomenclature remarks}
-Evolving networks are also called e.g. longitudinal networks, time-evolving networks, dynamic networks.
-We chose the term ``evolving network'' herein because it is simple, is semantically reasonable, and has been very much in use.
-The other main issue in current vocabulary is in the phrase ``sector of a network'' (e.g.
+\subsection{Remarks on nomenclature}
+Dynamic networks are also called e.g. longitudinal networks, time-evolving networks, evolving networks, and time-varying networks~\cite{ev,tev,tv,lg}.
+We chose to adopt the term ``dynamic network'' herein because it is simple, semantically meaningful and has been widely adopted.
+Another major issue in current vocabulary is in the phrase ``sector of a network'' (e.g.
 of the hubs).
 In most cases, the sectors are also partitions (i.e.
-they are non-overlapping sets whose union is the complete superset), but it depends of the criteria for obtaining the sectors~\cite{stab}, thus the use of the more general vocable \emph{sector}.
-Moreover, network science is very multidisciplinary and context-dependent, and flourished from the ill-defined area of complex systems, favoring polysemy and synonymy.
-We strove to avoid such issues because they are usually not convenient in scientific literature.
+they are non-overlapping sets whose union is the complete superset), but it depends on the criteria for obtaining the sectors~\cite{stab}, thus the use of the more general term \emph{sector}.
+Moreover, network science is highly multidisciplinary and context-dependent, and flourished from the ill-defined area of complex systems, favoring polysemy and synonymy.
+We strove to avoid such issues and keep a consistent usage of nomenclature.
+%because they are usually not convenient in scientific literature.
 
 \section{A description of the method}\label{sec:des}
-Versinus is essentially obtained through placing the nodes in segments
-with respect to their overall topological features, and then representing
+Versinus generates a layout through placing the nodes in segments
+with respect to their overall topological features and then representing
 instantaneous topological features though glyph and arrow attributes.
-The visualization may be divided thus in three stages, described in the following subsections.
+The visualization may be thus split in three stages, described in the following subsections.
 
 \subsection{Node placement}\label{sec:pla}
-First, a network is constructed with all the activity of interest, i.e. all the transactions, such as all the email messages considered or all the links in a data file.
-The nodes are then ordered either by degree or strength (or in- or out- degree or strength), from greatest to smallest values.
+First, a network is constructed with all the activity of interest, i.e. the transactions such as all the email messages considered, or all the links in a data file.
+The nodes are then ordered either by degree or strength (or in- or out-degree or strength), from greatest to smallest values.
 Also from such global network, the hubs, intermediary and peripheral sectors are obtained.
-Two methods have been proven useful for obtaining such sectors: 1) assuming fixed fractions of hubs (e.g.
+Two methods have been proven useful for obtaining sectors: 1) assuming fixed fractions of hubs (e.g.
 5\%), of intermediary (e.g.
 15\%) and peripheral nodes (e.g.
-80\%); 2) by comparison of the network against the Erdös-Rényi model, a non-trivial procedure, with relevant analytical results, carefully described in~\cite{stab}.
+80\%); 2) by comparison of the network against the Erdös-Rényi model, a non-trivial procedure, with relevant analytical results, detailed elsewhere~\cite{stab}.
 
-With such structures at hand, the nodes are placed along three segments.
+With such structures at hand, the nodes are placed along three segments, as depicted in Figure~\ref{fig:glob}.
 The hubs are placed along the first half of a sine period.
@@
 The peripheral nodes are placed along an independent line.
-Such placement is depicted in Figure~\ref{fig:glob} with
+The placement follows 
 the basic parameters:
@@
   \item The displacement $\Delta$ of the sine in the vertical axis.
-  \item the horizontal margin $\mu$, i.e. the distance from the left border to the first hubs, assumed the same as the distance from the right border to last intermediary node.
+  \item The horizontal margin $\mu$, i.e. the distance from the left border to the first hubs, assumed the same as the distance from the right border to last intermediary node.
   \item The positions $(x_0,y_0)$ and $(x_1,y_1)$, i.e. the endpoints of the segment for the peripheral sector.
@@
   \caption{The underlying layout of Versinus.
-Most connected nodes are positioned from left to right on the sine segment (for the hubs and intermediary sectors), and from right to left on the upper line (for the periphery sector).
+Highly connected nodes are positioned from left to right on the sine segment (for the hubs and intermediary sectors), and from right to left on the upper line (for the periphery sector).
 The parameters are: amplitude $\alpha$ of the oscillation; vertical displacement $\Delta$ of the oscillation; horizontal margin $\mu$; and line endpoints $(x_0, y_0)$ and $(x_1, y_1)$.
@@
 
-\noindent Some crafty notes: fine-tuning of the size of each sector (number of nodes) may come handy due to the density of nodes in each segment and due to their activity.
-Also, the node positioned at the middle of the sine period may be a hub or an intermediary node, and this choice often impacts final visualization clarity more than predicted by intuition.
-Finally, 
-avoiding clutter with the hubs motivated both 
+\noindent Some crafty remarks: fine-tuning the size (number of nodes) of each sector may come handy due to the node density and activity in each segment.
+Also, the node placed at the middle of the sine period may be a hub or an intermediary node, and this choice often impacts the clarity of the final visualization more than anticipated by intuition.
+Finally, avoiding clutter with the hubs motivated both 
 the positioning of the peripheral sector in the upper-right corner and
@@
 \subsection{Drawing nodes, links and auxiliary information}
-For visualizing the evolving network, further concepts and quantities need to be defined.
-A \emph{snapshot} $S$ consists of a number $\sigma$ time-contiguous transactions, e.g. of 100 email messages, starting at an arbitrary message.
+Additional concepts and quantities need to be defined for visualizing the evolving network as a data-driven animation.
+A \emph{snapshot} $S$ consists of a number $\sigma$ time-contiguous transactions, e.g. 100 email messages, starting at an arbitrary message.
 The separation $\eta$ between each snapshot is the number of transactions skipped from one snapshot to the next.
-In summary, the evolving network is visualized through a sliding window of $\sigma$ transactions, updated each $\eta$ transactions.
-To achieve the final animation, one need also to define a transaction rate $\omega$.
-Thus, these parameters are specified:
+In summary, the evolving network is visualized through a sliding window of $\sigma$ transactions, updated at every $\eta$ transactions.
+To achieve the final animation, a transaction rate $\omega$ must also be defined.
+Thus, the following parameters are specified:
 \begin{itemize}
   \item the window $\sigma$ captured at each snapshot, expressed in transactions.
-  \item The step size $\eta$ between each snapshot, expressed in transactions.
-  \item The transaction rate $\omega$, expressed in transactions per second.
+  \item the step size $\eta$ between each snapshot, expressed in transactions.
+  \item the transaction rate $\omega$, expressed in transactions per second.
 \end{itemize}
@@
 \noindent The sequence of snapshots may be expressed as $\{S_i^{\sigma, \eta}\}_0^{L}$, where $L$ is the smallest integer such that $L\times \eta + \sigma > T$ and $T$ is the total number of transactions.
-Notice that the last snapshot $S_L$ will not necessarily have $\sigma$ transactions, but may have less than $\sigma$.
-Also, the rate $\omega$ is only applied when rendering the animation.
-E.g.
+Notice that the final snapshot $S_L$ may have less than $\sigma$ transactions.
+Also, the rate $\omega$ is only applied when rendering the animation, e.g.
 if each snapshot $S_i$ yields an image, the stop motion animation will use $\omega/\eta$ images per second to achieve the desired rate of $\omega$ transactions per second.
 
-In each snapshot, the entailed network is analyzed in terms of the sectors (hubs, intermediary and periphery), in and out degree (or strength if proffered) and clustering coefficient.
-Then the visual mapping of information is performed in the following way:
+In each snapshot, the entailed network is analyzed in terms of the sectors (hubs, intermediary and periphery), in and out degree (or strength, if preferred) and clustering coefficient.
+Then the visual mapping of information is performed as follows:
 \begin{itemize}
-  \item nodes not incident in the snapshot are not shown. Nodes incident in the snapshot is given a minimum width and height.
-  \item Node sector in the snapshot is mapped to node shape (e.g. hexagon, circle and diamond for hub, intermediary and peripheral nodes).
-  \item Node width is incremented proportionally to its in-degree (or out-strength if preferred). Node height is incremented proportionally to its out-degree (or out-strength if preferred).
+  \item nodes not incident in the snapshot are not shown. Glyphs associated with nodes incident in the snapshot are assigned a minimum width and height.
+  \item The node sector in the snapshot is mapped to a glyph shape (e.g. circle, hexagon and diamond for hub, intermediary and peripheral nodes, respectively).
+  \item Node width is incremented proportionally to its in-degree (or out-strength, if preferred). Node height is incremented proportionally to its out-degree (or out-strength if preferred).
   \item Node color is related to clustering coefficient, e.g. zero is mapped to white, one is pure red, and intermediary values are scaled accordingly.
-  \item Link weight is mapped to link width and/or a color scale.
+  \item Link weight is mapped to line width and/or a color scale.
 \end{itemize}
 
-Finally, auxiliary information may be given by a persistent legend that displays essential information such as $\sigma$, $\eta$, $\omega$ and $T$, and keeps track of current snapshot along e.g.
-the number of nodes and edges of the snapshot.
-Also, node-specific information may be displayed by blinking corresponding cues over the nodes periodically or displaying them with a vertical displacement.
-Figure~\ref{fig:final} exemplifies two flavors of Versinus settings.
+\noindent Finally, complementary information may be displayed as a persistent legend that reveals essential information such as $\sigma$, $\eta$, $\omega$ and $T$, and keeps track of the current snapshot, e.g.
+its number of nodes and edges.
+Also, node-specific information may be displayed by blinking corresponding cues over the nodes periodically, displaying them with a vertical displacement from the node, or by writing them to a specific widget when requested by the analyst on an interactive interface, i.e.
+on-demand.
+Figure~\ref{fig:final} exemplifies two of the early flavors of Versinus settings.
 
 \begin{figure}[!h]\centering
-    \subfloat[Versinus flavour A]{{\includegraphics[width=.46\textwidth, height=4.3cm]{versinus_} }}%
-    \qquad
-  \subfloat[Versinus flavour B]{{\includegraphics[width=.46\textwidth, height=4.3cm]{im31} }}%
-    \caption{Two frames of animations achieved using the Versinus method.
-In (a) the individual degrees in the snapshot is blinking over the nodes, and the nodes are counted in auxiliary lines along the bottom and the near the periphery line.
-Current image/snapshot is 911 and the snapshot has 116 nodes and 262 links.
-   In (b), a cleaner version of Versinus is used to display a snapshot in which the only hub is the overall hub, the second overall hub is a peripheral node, and most nodes have high clustering coefficients.
-The legend indicates that the total number of transactions is $T=3998$, the window size $\sigma=40$, the step size is $\eta=40$, the transaction rate is $\omega=4$, current snapshot starts at transaction 1240, and the snapshot has 13 nodes and 15 links.
+    \subfloat[]{{\includegraphics[width=.46\textwidth, height=4.3cm]{versinus_} }}%
+    \quad
+  \subfloat[]{{\includegraphics[width=.46\textwidth, height=4.3cm]{im31} }}%
+    \caption{Two frames of dynamic network animations achieved using the Versinus method.
+    Both are achieved through dedicated Python scripts and package modules.
+    In (a) the individual degrees in the snapshot are blinking over the nodes, and the nodes are counted in auxiliary lines along the bottom and paralel to the periphery line.
+Current image/snapshot is 911 and has 116 nodes and 262 links.
+   In (b), a cleaner version is employed to display a snapshot in which the only hub is the overall hub, the second overall hub is a peripheral node, and most nodes have high clustering coefficient.
+The legend indicates that the total number of transactions is $T=3,998$, the window size $\sigma=40$, the step size is $\eta=40$, the transaction rate is $\omega=4$, the current snapshot starts at transaction 1,240 and has 13 nodes and 15 links.
+   The low-quality images are typical of the early implementations of Versinus, developed to assist the investigations of the researchers directly related to its development.
    Further information is given in Section~\ref{sec:des}.
@@
 Is the overall intermediary sector in greater activity with the hubs or periphery?
-Do higher clustering really occur most often in the less connected nodes?
+Does higher clustering really occur most often in the less connected nodes?
 Are there very unbalanced nodes with respect to in and out degree (or strength)?
-Are the number of nodes and edges stable along the network evolution?
+Is the number of nodes and edges stable along the network evolution?
 In fact, Versinus was partly designed to inspect network attributes due to the remarkable stability of their overall characteristics reported in~\cite{stab}.
-In this context, it is natural to ask if such stability is also noticeable in individual nodes, and, 
-as suggested by~\cite{barabasiEvo}, all hubs were found to have intermittent activity in social networks, except for the smallest networks analyzed.
+In this context, it is natural to ask if such stability is also noticeable in individual nodes or in connectivity patterns.
+Although the overall characteristics may be regarded as surprisingly stable, the stability of network components is very network-dependent.
+E.g.,  as suggested by~\cite{barabasiEvo}, all hubs were found to have intermittent activity in social networks, except for the smallest networks analyzed.
 
 \section{Software implementations}\label{sec:imp}
-There are four software implementations of Versinus.
-For organization and brevity, these are all linked in this repository:~\cite{verRepo}, together with this document.
-Two of them are written in Python: one is a script that renders animations as in the Figure~\ref{fig:final} (a), with more cues of local structure.
-The other is a library that renders animations as in the Figure~\ref{fig:final} (a), simpler and with more cues of the overall structure.
-The other two implementations are in JavaScript and uses WebGL.
-One is for static (non-evolving) networks, implemented as part of the ccNetViz library, that aims at being as computationally inexpensive as possible~\cite{cell}.
-The other is implemented in an under-development 3D Visual Analytics utility in which Versinus is combined with other analytical techniques, such as principal component analysis, overview first - focus on demand, and simultaneous highlighting of corresponding structures across images as required by user.
+Currently, there are four software implementations of Versinus known to the authors.
+For organization and brevity, these are all linked in\footnote{Pivot repository for Versinus resources such as implementations, public instances of interactive interfaces, and videos: https://github.com/ttm/versinus}.
+Two of them are written in Python: one is a script that renders animations as in the Figure~\ref{fig:final}
+(a), with more cues of local structure.
+The other is a library that renders animations as in the Figure~\ref{fig:final} (b), simpler and with more cues of the overall structure, and is capable of rendering sonifications of the dynamic networks, yielding ``audiovisualizations'', i.e.
+synced sonic and visual mappings of the dynamic networks.
+The other two implementations are in JavaScript and use WebGL.
+One is, by default, for static (non-dynamic) networks, implemented as part of the ccNetViz library, that aims at being as computationally inexpensive as possible~\cite{cell}.
+The other is an experimental 3D Visual Analytics application in which Versinus is combined with other analytical techniques,
+such as principal component analysis, overview first - focus on demand, and simultaneous highlighting of corresponding structures across images as required by users.
 These last two implementations are illustrated in Figure~\ref{fig:extra}.
@@
 \begin{figure}[!h]\centering
-    \subfloat[Versinus in JavaScript A]{{\includegraphics[width=.46\textwidth, height=2.5cm]{ccNetViz} }}%
+    \subfloat[Versinus in ccNetViz]{{\includegraphics[width=.46\textwidth, height=2.5cm]{ccNetViz} }}%
     \qquad
-  \subfloat[Versinus in JavaScript B]{{\includegraphics[width=.46\textwidth]{visAn} }}%
-    \caption{The two implementations of Versinus in JavaScript using WebGL.
-In (a) is an image of the implementation in ccNetViz, a library with the purpose of being computationally inexpensive.
-This implementation is only for static networks.
-In (b) is an image of the Versinus implementation in a Visual Analytics utility being developed.
-This latter software render the animations in real time, with multiple controls for inspecting the network structure and controlling the animation, in conjunction with other analysis techniques.
-Also, the sine oscillates both in the vertical and depth axes, as does the periphery segment.
+  \subfloat[Experimental Visual Analytics]{{\includegraphics[width=.46\textwidth]{visAn} }}%
+    \caption{Two implementations of Versinus in JavaScript using WebGL.
+(a) illustrates the implementation that is available within ccNetViz, a library purposed to being computationally inexpensive.
+This implementation handles static networks by default.
+(b) is an image achieved using an experimental Visual Analytics application.
+    It renders the animations in real time, with multiple controls for inspecting the network structure and controlling the animation, in association with other analysis techniques.
+The sine oscillates both in the vertical and depth axes, as does the periphery segment.
 Further information is given in Section~\ref{sec:imp}.}
@@
 The implementation illustrated in Figure~\ref{fig:final} (b) includes a sonification routine for the evolving network.
-The results were of very modest effectiveness in the sense that it does not enable the user to grasp network features with ease, specially if the user is not musically trained.
-Nevertheless, it made the (audio)visualization more interesting, and it is the first time the authors appreciated a sonification of an evolving network (maybe even of any network), and is thus probably a unique multimedia representation of networks.
-Accordingly, we believe valid a description of the sonification as it was left implemented, i.e.
-in its final version: the lower noise has central frequency proportional to first hub connectivity, at the extreme left on the layout, on the start of the sine, and it updated each two tempos.
-The higher noise has central frequency proportional to the second hub, and is also updated every two tempos.
-The third hub connectivity is proportional to the lower note, which is incident once a beat.
-The fourth hub connectivity is proportional to the higher note, one note occurring twice every half-beat.
+Results are of limited effectiveness in the sense that the sonification does not enable users to grasp network features with ease,
+specially if they are not musically trained.
+Nevertheless, it made the (audio)visualization more engaging, as it yields exotic media,
+and it is the first time the authors appreciated a sonification of an evolving network (maybe even of any network),
+and is thus probably a unique multimedia representation of networks.
+Accordingly, we understand suitable to provide a short description of the sonification as it was left implemented,
+i.e. in its final version,
+even though a full understanding of the summary requires knowledge of musical theory.
+Four sonic/musical voices are used, two of them are noises and the other two have defined pitch. The lower noise has central frequency proportional to first hub connectivity,
+which is the node positioned at the extreme left on the layout, on the starting point of the sine, and its frequency is updated each two tempos.
+The higher noise has central frequency proportional to the second hub, the node next to the first hub, and its frequency is also updated every two tempos.
+The third hub connectivity is proportional to the lower note,
+which is incident once a beat, and has its note value updated accordingly.
+The fourth hub connectivity is proportional to the higher note, which is incident twice every half-beat, is updated every note, and is only heard on the second half-beat of every tempo.
+Figure~\ref{fson} illustrates such disposition of sonic lines (or musical voices).
 
@@
 \section{Insights gathered}\label{sec:ins}
-Valuable insights were obtained by the design and use of Versinus.
-For example, the visualization of network structures are favored by the positioning of nodes in an oscillatory pattern.
+This section describes insights gathered by the authors,
+and should be further verified in subsequent work and using
+suitable validation methods.
+The legitimacy of the exposition herein given is based on the reasonable aspect of the assertions, and on the facts that the authors developed Versinus, used the method extensively, and the scientific contribution which it made possible~\cite{stab}.
+
+The visualization of network structures is favored by the placement of nodes in an oscillatory pattern, as it tends to decrease the superposition of nodes and links.
 In fact, this principle was implemented in some of the layouts available within ccNetViz, and it may be useful in other well-known layouts, such as the Hive Plot~\cite{hive}.
-The use of only one period for the sine is arbitrary and was established because it entailed good visualizations for the networks and tasks at hand during the design of Versinus, but is allowed to vary in more recent implementations.
+Using a single period for the sine is an arbitrary choice, established because it entailed good visualizations for the networks and tasks at hand during the design of Versinus.
+Nonetheless, it is allowed to vary in more recent implementations in JavaScript listed in~\footnote{Pivot repository for Versinus resources such as implementations, public instances of interactive interfaces, and videos: https://github.com/ttm/versinus}: the sine in which the hub and intermediary nodes are positioned may have as many periods as parametrized by the analyst, and the same is feature is available for the segment for the peripheral nodes.
 Another important insight is the pertinence of fixed positions for the observation of evolving networks.
-Not only this has the potential to help ameliorate the hairball effect~\cite{hairball}, but it is most often very difficult to keep track of individual nodes and structures when the nodes are allowed to move in the images, as in a force-based layout.
-The positioning of nodes with respect to their topological features (nodes are ordered by connectivity, segments are related to network sectors) revealed very useful for understanding network features, such as stability and connectivity dependent behaviors.
-Visual cues such as node width, height, shape and color were found informative for individual nodes but resulted in a cognitive overload for the observation of global features, which motivates simplified versions.
+Not only this has the potential to reduce the hairball effect~\cite{hairball},
+but it is most often very difficult to keep track of individual nodes and structures when the nodes are allowed to move in the images while the network evolves, as in force-based layouts.
+The placement of nodes with respect to their topological features (nodes are ordered by connectivity,
+segments are related to network sectors) revealed very useful for understanding network features,
+such as stability and connectivity-dependent behaviors.
+Visual cues such as glyph width, height, shape and color were found informative for grasping information on individual nodes,
+but results in a cognitive overload when attempting to observe global features,
+which motivates simplified settings.
 The sonification of the evolving network added to the exotic flavor of Versinus and made it more attractive.
-It seems that more informative sonifications will require a dedicated work, training of already musically competent users, and very modest goals.
+However, it seems that more informative sonifications should require dedicated research,
+training of already musically competent users, and very modest goals. 
 Furthermore, reasonable parametrization (e.g.
-through $\alpha$, $\Delta$, $\mu$, $(x_0, y_0)$, $(x_1, y_1)$, $\sigma$, $\eta$ and $\omega$) are very dependent of the network and intended tasks, and calls for interactive interfaces and automated indication of initial values.
+through $\alpha$, $\Delta$, $\mu$, $(x_0, y_0)$, $(x_1, y_1)$, $\sigma$, $\eta$ and $\omega$) is highly dependent on the network characteristics and user intended tasks, and calls for interactive interfaces and automated recommendation of initial values.
 
 \section{Conclusions and further work}\label{sec:con}
-Within this document, we achieved a first thorough description of the Versinus method, with its formal elements, use context and software implementations.
-The validity of Versinus is implied both by its use in published research scenarios and the insights it yielded on individual networks and network layouts in general.
-Further developments using the method are already being made, with focus on interactivity and Web technologies, such as WebGL and the use of well-polished JavaScript libraries.
+This document presents a description of the Versinus visualization method, with its fundamental elements, usage potentials, and software implementations.
+The validity of Versinus is implied both by its employment in published research~\cite{stab,thesis}, and the insights it yields on individual networks and network
+layouts in general.
+Further developments,
+with focus on interactivity and Web technologies,
+are currently being refined
+through research and software development.
 
@@
 valuable contribution, e.g. wiggling the segments of a Hive Plot~\cite{hive}.
-Most importantly, we aim at completing the Visual Analytics utility,
-enabling the use of Versinus in an interactive environment, for enhanced exploratory analysis.
-Finally, we plan to further analyze datasets of evolving networks~\cite{stab} using Versinus.
+We should implement further mechanisms for the detection of structural changes within the network evolution, such as using the persistent homology~\cite{hajij}.
+Most importantly, we aim at completing a Visual Analytics utility to support using Versinus in an interactive environment for enhanced exploratory data analysis,
+and verifying the claims described in Section~\ref{sec:ins} through formal validation techniques.
+Finally, we plan to further analyze data sets of evolving networks~\cite{stab} using Versinus.
 

### 2019-06-12 doc/paper.tex
adiantado artigo, backup

@@
 \keywords{Network visualization
-12
-
 \and Dynamic networks
@@
 \begin{itemize}
-  \item the amplitude $\alpha$ of the sine oscillation.
+  \item The amplitude $\alpha$ of the sine oscillation.
   \item The displacement $\Delta$ of the sine in the vertical axis.
@@
 \begin{itemize}
-  \item the window $\sigma$ captured at each snapshot, expressed in transactions.
-  \item the step size $\eta$ between each snapshot, expressed in transactions.
-  \item the transaction rate $\omega$, expressed in transactions per second.
+  \item The window $\sigma$ captured at each snapshot, expressed in transactions.
+  \item The step size $\eta$ between each snapshot, expressed in transactions.
+  \item The transaction rate $\omega$, expressed in transactions per second.
 \end{itemize}
@@
 \begin{itemize}
-  \item nodes not incident in the snapshot are not shown. Glyphs associated with nodes incident in the snapshot are assigned a minimum width and height.
+  \item Nodes not incident in the snapshot are not shown. Glyphs associated with nodes incident in the snapshot are assigned a minimum width and height.
   \item The node sector in the snapshot is mapped to a glyph shape (e.g. circle, hexagon and diamond for hub, intermediary and peripheral nodes, respectively).
@@
 
-\subsection{Network sonification}
+\subsection{Network sonification}\label{sson}
 The implementation illustrated in Figure~\ref{fig:final} (b) includes a sonification routine for the evolving network.
@@
 Four sonic/musical voices are used, two of them are noises and the other two have defined pitch. The lower noise has central frequency proportional to first hub connectivity,
-which is the node positioned at the extreme left on the layout, on the starting point of the sine, and its frequency is updated each two tempos.
-The higher noise has central frequency proportional to the second hub, the node next to the first hub, and its frequency is also updated every two tempos.
+which is the node positioned at the extreme left on the layout, on the starting point of the sine, and its frequency is updated every two beats.
+The higher noise has central frequency proportional to the second hub, the node next to the first hub, and its frequency is updated at each beat.
 The third hub connectivity is proportional to the lower note,
-which is incident once a beat, and has its note value updated accordingly.
-The fourth hub connectivity is proportional to the higher note, which is incident twice every half-beat, is updated every note, and is only heard on the second half-beat of every tempo.
+which is incident once a beat, and has its note value updated at every beat.
+The fourth hub connectivity is proportional to the higher note, which is incident twice every half-beat, is updated at every note, and is only heard on the second half-beat of every tempo.
 Figure~\ref{fson} illustrates such disposition of sonic lines (or musical voices).
 
+\begin{figure}[!h]\centering
+\includegraphics[width=.9\textwidth]{sonif}
+  \caption{Visual representation of the sonification implemented for the dynamic networks.
+  $c_f$ and $f_0$ stands for \emph{center frequency} and \emph{fundamental frequency}, respectively.
+  Further details are given in Section~\ref{sson}.
+  }\label{fson}
+\end{figure}
+
 \subsection{Animations produced and available}\label{sec:vid}
-Many animations were obtained through the routines described.
-These encompassed different networks and settings ($\alpha$, $\sigma$, $\eta$,
+A number of animations were obtained through the routines described.
+These encompassed different networks, settings ($\alpha$, $\sigma$, $\eta$,
 etc), and different flavors of Versinus.
-Some of these animations are available in six playlists of Youtube videos.
-For organization and brevity, they are available at~\cite{verRepo} together with other Versinus-related resources.
+Some of these visualizations are available in six playlists of Youtube videos.
+For organization and brevity, they are available at~\footnote{Pivot repository for Versinus resources such as implementations, public instances of interactive interfaces, and videos: https://github.com/ttm/versinus} together with other Versinus-related resources.
+
+\subsection{An experimental Visual Analytics interface}
+This section exemplifies the employment of the Versinus method
+within a more elaborate framework.
+Let us consider an interactive visual interface, with which the user may parametrize Versinus and then navigate through the entailed dynamic network.
+
+\begin{figure}[!h]\centering
+\includegraphics[width=\textwidth]{esfNetVis___}
+  \caption{
+  }\label{esfnetvis}
+\end{figure}
+
+\begin{figure}[!h]\centering
+\includegraphics[width=\textwidth]{esfNetVis2___}
+  \caption{
+  }\label{esfnetvis}
+\end{figure}
+
+\begin{figure}[!h]\centering
+\includegraphics[width=\textwidth]{esfNetVis3}
+  \caption{
+  }\label{esfnetvis}
+\end{figure}
+
+\begin{figure}[!h]\centering
+\includegraphics[width=\textwidth]{esfNetVis4}
+  \caption{
+  }\label{esfnetvis}
+\end{figure}
+One basic feature of the method is the constant representation of
+nodes in each of the hubs, intermediary and peripheral sectors.
+Nodes are positioned according to such classification regarding all
+the activity of interest, and in each snapshot visual cues map the
+classification within the snapshot.
+Following the analytic methods used in~\cite{stab}, it may be crucial
+to know how the relative sizes of the sectors vary along time, i.e. along
+the sequence of snapshots.
 
@@
 Using a single period for the sine is an arbitrary choice, established because it entailed good visualizations for the networks and tasks at hand during the design of Versinus.
-Nonetheless, it is allowed to vary in more recent implementations in JavaScript listed in~\footnote{Pivot repository for Versinus resources such as implementations, public instances of interactive interfaces, and videos: https://github.com/ttm/versinus}: the sine in which the hub and intermediary nodes are positioned may have as many periods as parametrized by the analyst, and the same is feature is available for the segment for the peripheral nodes.
+Nonetheless, it is allowed to vary in more recent implementations in JavaScript\footnote{Pivot repository for Versinus resources such as implementations, public instances of interactive interfaces, and videos: https://github.com/ttm/versinus}: the sine in which the hub and intermediary nodes are positioned may have as many periods as parametrized by the analyst, and the same is feature is available for the segment for the peripheral nodes.
 Another important insight is the pertinence of fixed positions for the observation of evolving networks.

### 2019-06-12 doc/mkFigBr.py

@@ -1,0 +1,47 @@
+import pylab as p, numpy as n
+
+fig=p.figure(figsize=(6.,4.))
+p.xlim(0,1)
+p.ylim(0,1)
+
+xmargin = 0.1
+amp = 0.7
+offset = 0.1
+x = n.linspace(0+xmargin, 0.5, 1000)
+y = 0.5 + 0.5 * amp * n.sin(n.linspace(0, n.pi, 1000)) - offset
+p.plot(x, y, label='hubs')
+
+p.annotate('', (1-xmargin, 0.2), (1, 0.2), ha="center", va="center", arrowprops=dict(arrowstyle="|-|"))
+p.annotate(r'$\mu$', (0.94, 0.22), fontsize=14)
+
+# p.arrow(0.1, 0.7, 0.1, 0.2, arrowprops=dict(arrowstyle="|-|"))
+p.annotate('', (0.05, y.max()), (0.05, y.min()), ha="center", va="center", arrowprops=dict(arrowstyle="|-|"))
+p.annotate(r'$\alpha$', (0.07, y.max()*0.75), fontsize=14)
+
+p.annotate('', (0.2, 0.5), (0.2, y.min()), ha="center", va="center", arrowprops=dict(arrowstyle="|-|"))
+p.annotate(r'$\Delta$', (0.22, 0.43), fontsize=14)
+
+x = n.linspace(0.5, 1-xmargin, 1000)
+y = 0.5 + 0.5 * amp * n.sin(n.linspace(n.pi, 2*n.pi, 1000)) - offset
+p.plot(x, y, label='intermediarios')
+
+x0, y0 = 0.8, 0.6
+x1, y1 = 0.4, 0.9
+
+xx = n.linspace(x0, x1, 100)
+yy = n.linspace(y0, y1, 100)
+ax = p.plot(xx, yy, label='periféricos')
+
+p.annotate(r'$(x_0, y_0)$', (x0, y0+.02), fontsize=14)
+p.annotate(r'$(x_1, y_1)$', (x1-0.02, y1+0.02), fontsize=14)
+
+p.xticks([0, 0.5, 1])
+p.yticks([0, 0.5, 1])
+p.legend()
+
+p.title("Posicionamento dos vértices no Versinus", fontsize=16)
+
+p.subplots_adjust(left=0.06, bottom=0.06, right=0.97, top=0.93)
+
+p.savefig("./nodePositioningBr.png")
+# p.show()

### 2019-06-12 doc/mkSonifFig.py

@@ -1,0 +1,133 @@
+import pylab as p, numpy as n
+
+fig=p.figure(figsize=(6.,4.))
+p.xlim(-0.5,12)
+p.ylim(-0.5,12)
+
+p.ylabel(r'frequency ($Herz$) $\rightarrow$')
+p.xlabel(r'time ($beats$) $\rightarrow$')
+p.yticks([0],[''])
+
+xy = [
+    [0,1.9],
+    [2,3.9],
+    [4,5.9],
+    [6,7.9],
+    [8,9.9],
+    ],[
+    [1,1],
+    [0,0],
+    [2,2],
+    [1.3,1.3],
+    [0.6,0.6],
+]
+for x, y in zip(*xy):
+    p.plot(x, y, 'r',)
+p.plot([],[],'r',label=r'$c_f$ of hub noise')
+
+xy = [
+    [0,0.9],
+    [1,1.9],
+    [2,2.9],
+    [3,3.9],
+    [4,4.9],
+    [5,5.9],
+    [6,6.9],
+    [7,7.9],
+    [8,8.9],
+    [9,9.9],
+    ],[
+    [2.7,2.7],
+    [2.5,2.5],
+    [3,3],
+    [3,3],
+    [2.6,2.6],
+    [3,3],
+    [2.3,2.3],
+    [3.3,3.3],
+    [3.6,3.6],
+    [2.6,2.6],
+]
+for x, y in zip(*xy):
+    p.plot(x, y, 'g',)
+p.plot([],[],'g',label=r'$c_f$ of second hub noise')
+
+xy = [
+    [0,0.9],
+    [1,1.9],
+    [2,2.9],
+    [3,3.9],
+    [4,4.9],
+    [5,5.9],
+    [6,6.9],
+    [7,7.9],
+    [8,8.9],
+    [9,9.9],
+    ],[
+    [4.7,4.7],
+    [3.7,3.7],
+    [4,4],
+    [5,5],
+    [4.6,4.6],
+    [4,4],
+    [4.3,4.3],
+    [5.3,5.3],
+    [4.6,4.6],
+    [3.9,3.9],
+]
+for x, y in zip(*xy):
+    p.plot(x, y, 'b',)
+p.plot([],[],'b',label=r'$f_0$ of third hub note')
+
+xy = [
+        [0.5, 0.73],
+        [0.75,0.98],
+    [1.5,1.73],
+    [1.75,1.98],
+        [2.5, 2.67],
+        [2.80,2.98],
[…]

### 2019-06-13 doc/paper.tex
primeira versao complea, falta lapidar

@@
 \begin{figure}[!h]\centering
-\includegraphics[width=.9\textwidth]{sonif}
+\includegraphics[width=0.7\textwidth]{sonif}
   \caption{Visual representation of the sonification implemented for the dynamic networks.
@@
 
-\subsection{An experimental Visual Analytics interface}
-This section exemplifies the employment of the Versinus method
+\subsection{An experimental visual analytics interface}
+This section exemplifies the employment of the Versinus network visualization method
 within a more elaborate framework.
 Let us consider an interactive visual interface, with which the user may parametrize Versinus and then navigate through the entailed dynamic network.
+A most fundamental Versinus parametrization is performed before the visual mapping of the network, and is illustrated in Figure~\ref{env1}.
+Initial visualization is illustrated in Figure~\ref{env2}.
+Figures~\ref{env3} and~\ref{env4} exemplifies Versinus compliant interactive data visualization analytical procedures: the observation of the fraction of nodes in each sector, characteristics of individual nodes and of snapshots.
+The interactive interface is named $ESFNetVis$ (after ``Evolutive Scale Free Network Visualization'').
+At\footnote{Pivot repository for Versinus resources such as implementations, public instances of interactive interfaces, and videos: https://github.com/ttm/versinus} are directions for the code and a publicly accessible instance of $ESFNetVis$.
 
 \begin{figure}[!h]\centering
-\includegraphics[width=\textwidth]{esfNetVis___}
-  \caption{
-  }\label{esfnetvis}
+\includegraphics[width=0.7\textwidth]{esfNetVis___}
+  \caption{Visual elements for settings performed previous to the visual mapping of Versinus.
+  The buttons on the top are for keeping, modifying and accessing settings.
+  The Widgets in the gray rectangles are for selecting the network, selecting the first and last messages (transactions) to be considered,
+  and the sectorialization method.
+  The widgets in the white rectangle are $\sigma$, $\eta$, and $\omega$.
+  These settings are compliant with the description available in Section~\ref{sec:des}.
+  When the red button is clicked, the dynamic network is mapped to the canvas and additional elements, as described in Figure~\ref{env2}.
+  }\label{env1}
 \end{figure}
@@
 \begin{figure}[!h]\centering
-\includegraphics[width=\textwidth]{esfNetVis2___}
-  \caption{
-  }\label{esfnetvis}
+\includegraphics[width=0.7\textwidth]{esfNetVis2___}
+  \caption{The initial visualization of the dynamic network performed using $ESFNetVis$.
+  In the canvas, the network is mapping using a 3D extension of Versinus, in which both segments oscillates vertically and in depth.
+  The red portions of the links denotes the direction of the link (often performed using arrow heads).
+  The text area starts with overall information of the collection of snapshots (e.g.
+each snapshot yields a mean degree, the set of snapshots yields a mean of such mean degree and a standard deviation of the mean degree.
+  The timelines hold the fractions of nodes in each sector with orange, green and blue for hub, intermediary and peripheral nodes.
+  The red bar is the playback cue, i.e. the snapshot being visualized.
+  The analyst may play and pause with the play button,
+  and rewind with the rewind button, both to the right.
+  }\label{env2}
 \end{figure}
@@
 \begin{figure}[!h]\centering
-\includegraphics[width=\textwidth]{esfNetVis3}
-  \caption{
-  }\label{esfnetvis}
+\includegraphics[width=0.7\textwidth]{esfNetVis3}
+  \caption{The layout is slightly rotated in order to focus on the peripheral nodes segment.
+This image depicts one of the interaction possibilities among the canvas and the timelines: when the mouse is over a node, the line of the sector of the node in the snapshot is made black.
+  (Notice the pointer in ``button-mode'', i.e. visualized as a hand.)
+  Also, the text area in this image shows information about specific nodes, provided on-demand (the user hits the key ``i'' when the mouse is over a node for the interface to provide the information).
+  }\label{env3}
 \end{figure}
@@
 \begin{figure}[!h]\centering
-\includegraphics[width=\textwidth]{esfNetVis4}
+\includegraphics[width=0.7\textwidth]{esfNetVis4}
   \caption{
-  }\label{esfnetvis}
+  The layout is rotated in order to focus on the intermediary segment of the sinusoid.
+  The text area shows information about specific snapshots, given on-demand. The user requests the information by pressing "S" while the snapshot is being visualized.
+  the nodes in the intermediary sectors are being highlighted on-demand by the user: notice the pointer which is over the intermediary sector timeline.
+  }\label{env4}
 \end{figure}

### 2019-06-13 doc/paper.tex
todos os arquivos usados

--- 2019-06-12 doc/paper.tex
@@
 %
-\title{Versinus: an Animated Visualization Method for Dynamic Scale-Free Networks\thanks{Supported by FAPESP, project 2017/05838-3}}
+\title{An Animated Visualization Method for Dynamic Scale-Free Networks\thanks{Supported by FAPESP, project 2017/05838-3}}
 %
@@
 There is a growing interest in modeling dynamic (or time-evolving, or longitudinal) networks and
-a number of procedures have been proposed to visualize them.
+a number of procedures have been proposed to enable their visualization.
   This article presents a visualization method for dynamic scale-free networks.
-  Compliant methods were not found in other tools by the authors, which motivated this description and the development of a dedicated interface.
-  Named Versinus, the visualization is specially useful in the observation of (in)stability of basic topological characteristics, and consists essentially in placing the most connected nodes (hubs and intermediary) along a sine curve, and the peripheral nodes along a separate segment, usually a line.
-  Thus its name, from  Latin \emph{versus} (line) and \emph{sinus} (sine).
-  The Versinus visualization can provide insight into network properties and is useful for observing its behavior, as described in this paper, after a formal presentation of the method and its associated software implementations.
+  Named Versinus, the method consists of a specialized combination of layout and further visual cues
+  useful in the observation of basic topological characteristics.
+  The layout consists essentially in placing the most connected nodes (hubs and intermediary)
+  along a sine curve, and the peripheral nodes along a separate segment, usually a line,
+  thus its name, from the Latin \emph{versus} (line) and \emph{sinus} (sine).
+  The visual cues yield evident both local and global key-features.
+  A compliant method was not found in the scientific literature or among existing tools by the authors.
+  Therefore, after a formal presentation of Versinus and its associated software implementations,
+  a simple case study is described in order to provide insights into the potential exploitation of the method
+  in analysing dynamic scale-free networks.
+  % which motivated this description and the development of a dedicated interface.
 \keywords{Network visualization
+\and Scale-free networks
 \and Dynamic networks
-\and Longitudinal networks
-\and Time-evolving networks
-\and Evolving networks
 \and Animated visualization
@@
 \end{abstract}
-%
-%
-%
+
 \section{Introduction}
-The visualization of dynamic networks poses challenges related to data size and complexity which have merited various contributions~\cite{evo1,hajij,evo3,evo2}.
-In fact, the meaningful visualization of dynamic networks is a well-known problem in Information Visualization and dynamic network analysis.
-In this work, we present an approach for dynamic network visualization that consists in distributing the nodes along a fixed layout arranged in three segments associated with the three sectors (or partitions) of a scale-free network, namely for the hub, intermediary and peripheral nodes~\cite{stab}.
-The network dynamic features are expressed by means of changes in visual properties of the glyphs used to depict nodes and links, e.g., glyph height, width, color and shape, and line (arrow) disposition, thickness and color.
-Moreover, complementary global and local information is rendered in order to enable an acoustic depiction of features of interest.
-The method has been called Versinus since its early usages in 2013, because the layout stabilized, after attempts with distinct arrangements, in a sine period and a line, which in Latin is \emph{sinus} (sine) and \emph{versus} (line).
-There are four software implementations of Versinus, two of them written in Python scripts and packages that yield animations through stop-motion.
-The other two are written in JavaScript, one focused on exploitation of static networks, the other is an experimental Visual Analytics interface, which extends the visual metaphor to the 3D space and provides additional analysis techniques such as principal component analysis (PCA) and resources for structure navigation\footnote{An instance of the interface is available at: http://rfabbri.vicg.icmc.usp.br:3000/evolution}.
-The writing of this document was motivated by
-the fact that no method was found by the authors that aims at visualizing dynamic scale-free networks through node-link diagrams.
-Furthermore, the visualization method was employed in scientific research~\cite{thesis,stab}, and thus a careful mathematical and technological description of the procedures involved in applying Versinus to real networks is appropriate.
+The visualization of dynamic networks poses challenges related to data size and
+complexity which have merited various
+contributions~\cite{evo1,evo3,hajij,evo2}.  In fact, the meaningful
+visualization of dynamic networks is a well-known problem in Information
+Visualization and dynamic network analysis.
+In this work, we present a peculiar
+method for dynamic network visualization which consists in distributing the
+nodes along a fixed layout and mapping the network evolution through changes in
+glyphs corresponding to nodes and links.
+The layout is arranged in three segments, each associated with one of the three
+sectors (or partitions) of a scale-free network, namely the hub,
+intermediary and peripheral nodes~\cite{stab}.
+Local features of the network are mapped to the visual properties of the glyphs,
+e.g., glyph height, width, color and shape, and line (arrow)
+thickness and color. Moreover, complementary global and local
+information is rendered in order to enable the observation of basic aspects
+of the dynamic network.
+The method has been called Versinus since its early usages in 2013,
+because the layout settled, after attempts with distinct arrangements, in a
+sine period and a line, which in Latin is \emph{sinus} (sine) and \emph{versus}
+(line).
+No compliant method was found by the authors
+for visualizing dynamic scale-free networks through node-link
+diagrams.  Furthermore, the visualization method was employed in scientific
+research~\cite{thesis,stab} and has received a number of software implementations.
+
+% There are four software implementations of Versinus, two of them
+% written through Python scripts and packages, and yield animations through
+% stop-motion. The other two are written in JavaScript, one is focused on the
+% exploitation of static networks, the other is an experimental Visual Analytics
+% interface, which extends the visual metaphor to the 3D space and provides
+% additional analysis techniques such as principal component analysis (PCA) and
+% resources for structure navigation\footnote{Central repository for Versinus resources such as implementations, public instances of interactive interfaces, and videos: https://github.com/ttm/versinus}.
+% The writing of
 
 The article is organized as follows.
-The next subsections address related work and issues of nomenclature.
-Section~\ref{sec:des} describes the method in detail and sufficient formalism.
-Section~\ref{sec:imp} outlines the software implementations of Versinus and depicts the resulting animations.
-Section~\ref{sec:ins} is dedicated to revealing further insights obtained through the use of the method.
+The next subsections address related work and nomenclature.
+Section~\ref{sec:des} describes the method.
+Section~\ref{sec:imp} outlines the software implementations of Versinus and depicts the resulting animations and interfaces.
+Section~\ref{visAn} describes a simple case study in order to provide insights about the potentialities of the method.
+Section~\ref{sec:ins} is dedicated to hypotheses provided through the use of the method.
 Finally, Section~\ref{sec:con} assembles concluding commentaries and further work.
@@
 % on vis of evo nets
-Although animation for the purposes of data analysis is often criticized in the data visualization literature, advantages have also been reported~\cite{anim,munzner,cog,ware}, e.g.: yielding visualizations more eye catching and taking advantage of the fact that humans are used to interacting with moving objects.
+Although animation for the purposes of data analysis is often criticized in the data visualization literature, advantages have also been reported~\cite{anim,munzner,cog,ware}, e.g.: for eye catching visualizations and potential advantages yield by the fact that humans are used to interacting with moving objects.
 Dynamic network visualization has been tackled both through animated and static (often timeline-based) visualizations~\cite{hajij,brain,visAn,ego}, focusing on various global and local network features.
-The contribution reported in this document is singular in that it consists basically of a layout which is fit for networks with scale-free characteristics, in association with some visualization cues which are implementation dependent.
+The contribution reported in this document is possibly singular in that it consists basically of a layout which was designed for networks with scale-free characteristics, in association with some visual cues.
 It results from extensive attempts to reach a satisfactory visualization by means theoretical and intuitive strategies and methodical verification of the result by appreciation of the entailed animations.
-The procedures and conceptualizations involved in achieving the final visualization were also considered worth documenting, as they were not found elsewhere in the literature, and by virtue of the simple and genuine design.
+% The procedures and conceptualizations involved in achieving the final visualization were also considered worth documenting, as they were not found elsewhere in the literature, and by virtue of the simple and genuine design.
 Its scientific relevance on a practical application has been demonstrated before~\cite{stab}.
@@
 Dynamic networks are also called e.g. longitudinal networks, time-evolving networks, evolving networks, and time-varying networks~\cite{ev,tev,tv,lg}.
-We chose to adopt the term ``dynamic network'' herein because it is simple, semantically meaningful and has been widely adopted.
-Another major issue in current vocabulary is in the phrase ``sector of a network'' (e.g.
-of the hubs).
+We adopted the term ``dynamic network'' herein because it is simple, semantically meaningful and has been widely adopted.
+Another major issue in current vocabulary arises when considered the hub, intermediary and peripheral sectors.
 In most cases, the sectors are also partitions (i.e.
 they are non-overlapping sets whose union is the complete superset), but it depends on the criteria for obtaining the sectors~\cite{stab}, thus the use of the more general term \emph{sector}.
-Moreover, network science is highly multidisciplinary and context-dependent, and flourished from the ill-defined area of complex systems, favoring polysemy and synonymy.
-We strove to avoid such issues and keep a consistent usage of nomenclature.
+Moreover, network science is highly multidisciplinary and context-dependent, and flourished from the vaguely-defined area of complex systems, favoring polysemy and synonymy.
+We strove to avoid such issues and to employ a consistent.
 %because they are usually not convenient in scientific literature.
@@
 \section{A description of the method}\label{sec:des}
-Versinus generates a layout through placing the nodes in segments
+Versinus layout consists basically in placing the nodes in segments
 with respect to their overall topological features and then representing
-instantaneous topological features though glyph and arrow attributes.
-The visualization may be thus split in three stages, described in the following subsections.
+instantaneous topological features through glyph attributes.
+Additional, implementation-dependent, information is often provided.
+The achievement of the visualization may be thus split in three stages,
+described in the following subsections.
 
@@
 Also from such global network, the hubs, intermediary and peripheral sectors are obtained.
-Two methods have been proven useful for obtaining sectors: 1) assuming fixed fractions of hubs (e.g.
+Two methods have proven useful for obtaining the sectors: 1) assuming fixed fractions of hubs (e.g.
 5\%), of intermediary (e.g.
 15\%) and peripheral nodes (e.g.
-80\%); 2) by comparison of the network against the Erdös-Rényi model, a non-trivial procedure, with relevant analytical results, detailed elsewhere~\cite{stab}.
+80\%); 2) comparison of the network against the Erdös-Rényi model, a non-trivial procedure, with relevant analytical results, detailed in~\cite{stab}.
 
@@
 The peripheral nodes are placed along an independent line.
-The placement follows 
-the basic parameters:
+These are the basic parameters:
 \begin{itemize}
@@
   \item The displacement $\Delta$ of the sine in the vertical axis.
-  \item The horizontal margin $\mu$, i.e. the distance from the left border to the first hubs, assumed the same as the distance from the right border to last intermediary node.
+  \item The horizontal margin $\mu$, i.e. the distance from the left border to the first hub, assumed to be the same as the distance from the right border to last intermediary node.
   \item The positions $(x_0,y_0)$ and $(x_1,y_1)$, i.e. the endpoints of the segment for the peripheral sector.
@@
 The parameters are: amplitude $\alpha$ of the oscillation; vertical displacement $\Delta$ of the oscillation; horizontal margin $\mu$; and line endpoints $(x_0, y_0)$ and $(x_1, y_1)$.
-   Further information is given in Section~\ref{sec:pla}.
+   Further information is provided in Section~\ref{sec:pla}.
   }\label{fig:glob}
@@
 
-\noindent Some crafty remarks: fine-tuning the size (number of nodes) of each sector may come handy due to the node density and activity in each segment.
-Also, the node placed at the middle of the sine period may be a hub or an intermediary node, and this choice often impacts the clarity of the final visualization more than anticipated by intuition.
+\noindent Remarks: fine-tuning the size (number of nodes) of each sector may come handy due to the node density and activity in each segment.
+Also, the node placed at the middle of the sine period may be a hub or an intermediary node, and this choice often impacts the clarity of the final visualization more than anticipated by intuition, specially in small networks.
 Finally, avoiding clutter with the hubs motivated both 
@@
 \subsection{Drawing nodes, links and auxiliary information}
-Additional concepts and quantities need to be defined for visualizing the evolving network as a data-driven animation.
-A \emph{snapshot} $S$ consists of a number $\sigma$ time-contiguous transactions, e.g. 100 email messages, starting at an arbitrary message.
+Let a \emph{snapshot} $S$ consist of a number $\sigma$ time-contiguous transactions, e.g. 100 email messages, starting at an arbitrary transaction.
 The separation $\eta$ between each snapshot is the number of transactions skipped from one snapshot to the next.
-In summary, the evolving network is visualized through a sliding window of $\sigma$ transactions, updated at every $\eta$ transactions.
+The evolving network is thus conceptualized as a sliding window of $\sigma$ transactions, which progresses by skipping $\eta$ transactions at each step.
 To achieve the final animation, a transaction rate $\omega$ must also be defined.
-Thus, the following parameters are specified:
+Thus, the following parameters are defined:
 \begin{itemize}
-  \item The window $\sigma$ captured at each snapshot, expressed in transactions.
+  \item The window size of $\sigma$ captured at each snapshot, expressed in transactions.
   \item The step size $\eta$ between each snapshot, expressed in transactions.
@@
 if each snapshot $S_i$ yields an image, the stop motion animation will use $\omega/\eta$ images per second to achieve the desired rate of $\omega$ transactions per second.
+In other words, the transactions in each snapshot are defined by $\sigma$ and $\eta$ and is independent from $\omega$.
 
@@
 
-\noindent Finally, complementary information may be displayed as a persistent legend that reveals essential information such as $\sigma$, $\eta$, $\omega$ and $T$, and keeps track of the current snapshot, e.g.
+\noindent Finally, complementary information may be displayed as a persistent legend that reveals essential information such as $\sigma$, $\eta$, $\omega$ and $T$, and other information that assists the analyst in keeping track of essential features of the current snapshot, e.g.
 its number of nodes and edges.
@@
 on-demand.
-Figure~\ref{fig:final} exemplifies two of the early flavors of Versinus settings.
+Figure~\ref{fig:final} exemplifies two early Versinus settings.
 
@@
     \caption{Two frames of dynamic network animations achieved using the Versinus method.
-    Both are achieved through dedicated Python scripts and package modules.
-    In (a) the individual degrees in the snapshot are blinking over the nodes, and the nodes are counted in auxiliary lines along the bottom and paralel to the periphery line.
-Current image/snapshot is 911 and has 116 nodes and 262 links.
+    In (a) the individual degrees in the snapshot are blinking over the nodes, and the nodes are counted in auxiliary lines along the bottom and parallel to the periphery line.
+Current image/snapshot is 911, with 116 nodes and 262 links.
    In (b), a cleaner version is employed to display a snapshot in which the only hub is the overall hub, the second overall hub is a peripheral node, and most nodes have high clustering coefficient.
-The legend indicates that the total number of transactions is $T=3,998$, the window size $\sigma=40$, the step size is $\eta=40$, the transaction rate is $\omega=4$, the current snapshot starts at transaction 1,240 and has 13 nodes and 15 links.
+The legend indicates that the total number of transactions is $T=3,998$, the window size is $\sigma=40$, the step size is $\eta=40$, the transaction rate is $\omega=4$, the current snapshot starts at transaction 1,240, and has 13 nodes and 15 links.
    The low-quality images are typical of the early implementations of Versinus, developed to assist the investigations of the researchers directly related to its development.
-   Further information is given in Section~\ref{sec:des}.
+   Further information is provided in Section~\ref{sec:des}.
    }%
@@
 Is the overall intermediary sector in greater activity with the hubs or periphery?
-Does higher clustering really occur most often in the less connected nodes?
+Does higher clustering occur most often in the less connected nodes?
 Are there very unbalanced nodes with respect to in and out degree (or strength)?
@@
 In fact, Versinus was partly designed to inspect network attributes due to the remarkable stability of their overall characteristics reported in~\cite{stab}.
-In this context, it is natural to ask if such stability is also noticeable in individual nodes or in connectivity patterns.
+In this context, it is natural to ask if such stability is also noticeable in individual nodes or in the connectivity patterns.
 Although the overall characteristics may be regarded as surprisingly stable, the stability of network components is very network-dependent.
-E.g.,  as suggested by~\cite{barabasiEvo}, all hubs were found to have intermittent activity in social networks, except for the smallest networks analyzed.
+E.g., as reported in~\cite{barabasiEvo}, hubs exhibit intermittent activity in social networks, except for the smallest networks analyzed.
 
@@
 Currently, there are four software implementations of Versinus known to the authors.
-For organization and brevity, these are all linked in\footnote{Pivot repository for Versinus resources such as implementations, public instances of interactive interfaces, and videos: https://github.com/ttm/versinus}.
-Two of them are written in Python: one is a script that renders animations as in the Figure~\ref{fig:final}
-(a), with more cues of local structure.
-The other is a library that renders animations as in the Figure~\ref{fig:final} (b), simpler and with more cues of the overall structure, and is capable of rendering sonifications of the dynamic networks, yielding ``audiovisualizations'', i.e.
-synced sonic and visual mappings of the dynamic networks.
-The other two implementations are in JavaScript and use WebGL.
+For organization and brevity, these are all linked in\footnote{Central repository for Versinus resources such as implementations, public instances of interactive interfaces, and videos: https://github.com/ttm/versinus}.
+Two of them are written in Python: one is a script that renders animations as in the Figure~\ref{fig:final}~(a),
+with more cues of local structure.
+The other is a library that renders animations as in the Figure~\ref{fig:final}~(b), simpler and with more cues of the overall structure, and is capable of rendering sonifications of the dynamic networks, yielding ``audiovisualizations'', i.e.
+synced sonic and visual mappings of the dynamic networks, as described in Section~\ref{sson}.
+The other two implementations are achieved through JavaScript and WebGL.
 One is, by default, for static (non-dynamic) networks, implemented as part of the ccNetViz library, that aims at being as computationally inexpensive as possible~\cite{cell}.
-The other is an experimental 3D Visual Analytics application in which Versinus is combined with other analytical techniques,
+The other is an experimental interactive visualization application in which Versinus is combined with other analytical techniques,
 such as principal component analysis, overview first - focus on demand, and simultaneous highlighting of corresponding structures across images as required by users.
@@
     \qquad
-  \subfloat[Experimental Visual Analytics]{{\includegraphics[width=.46\textwidth]{visAn} }}%
+  \subfloat[Experimental interactive visualization]{{\includegraphics[width=.46\textwidth]{visAn} }}%
     \caption{Two implementations of Versinus in JavaScript using WebGL.
@@
 This implementation handles static networks by default.
-(b) is an image achieved using an experimental Visual Analytics application.
-    It renders the animations in real time, with multiple controls for inspecting the network structure and controlling the animation, in association with other analysis techniques.
-The sine oscillates both in the vertical and depth axes, as does the periphery segment.
-Further information is given in Section~\ref{sec:imp}.}
+(b) is an image achieved using an experimental interactive visualization application.
+    This implementation renders the animations in real time, with multiple controls for inspecting the network structure and controlling the animation, in association with other analysis techniques.
+The sine oscillates both in the vertical and depth axes, as does the periphery segment.}
     \label{fig:extra}
@@
 Nevertheless, it made the (audio)visualization more engaging, as it yields exotic media,
-and it is the first time the authors appreciated a sonification of an evolving network (maybe even of any network),
-and is thus probably a unique multimedia representation of networks.
-Accordingly, we understand suitable to provide a short description of the sonification as it was left implemented,
-i.e. in its final version,
-even though a full understanding of the summary requires knowledge of musical theory.
+and it is the first time the authors appreciated a sonification of an evolving network (maybe even of any network).
+It is possibly a unique multimedia representation of networks.
+Accordingly, we understand suitable to provide a short description of the sonification as it was left implemented, i.e. in its final version,
+even though a full understanding of the summary requires knowledge of musical theory~\cite{mass}.
 Four sonic/musical voices are used, two of them are noises and the other two have defined pitch. The lower noise has central frequency proportional to first hub connectivity,
@@
 \begin{figure}[!h]\centering
-\includegraphics[width=.9\textwidth]{sonif}
+\includegraphics[width=0.7\textwidth]{sonif}
   \caption{Visual representation of the sonification implemented for the dynamic networks.
   $c_f$ and $f_0$ stands for \emph{center frequency} and \emph{fundamental frequency}, respectively.
-  Further details are given in Section~\ref{sson}.
+  Further information is provided in Section~\ref{sson}.
   }\label{fson}
@@
 A number of animations were obtained through the routines described.
-These encompassed different networks, settings ($\alpha$, $\sigma$, $\eta$,
+They encompass different networks, settings ($\alpha$, $\sigma$, $\eta$,
 etc), and different flavors of Versinus.
-Some of these visualizations are available in six playlists of Youtube videos.
-For organization and brevity, they are available at~\footnote{Pivot repository for Versinus resources such as implementations, public instances of interactive interfaces, and videos: https://github.com/ttm/versinus} together with other Versinus-related resources.
+For organization and brevity, they are available at~\footnote{Central repository for Versinus resources such as implementations, public instances of interactive interfaces, and videos: https://github.com/ttm/versinus}.
 
-\subsection{An experimental Visual Analytics interface}
-This section exemplifies the employment of the Versinus method
-within a more elaborate framework.
+\section{A simple case study}\label{visAn}
+This section exemplifies the employment of the Versinus network visualization method
+within a more elaborate framework through a very simple case study.
 Let us consider an interactive visual interface, with which the user may parametrize Versinus and then navigate through the entailed dynamic network.
+A most fundamental Versinus parametrization is performed before the visual mapping of the network, and is illustrated in Figure~\ref{env1}.
+Initial visualization is illustrated in Figure~\ref{env2}.
+Figures~\ref{env3} and~\ref{env4} exemplifies Versinus compliant interactive data visualization procedures: the observation of the nodes in each sector, and characteristics of individual nodes and of snapshots.
+The interactive interface is named $ESFNetVis$ (after ``Evolutive Scale Free Network Visualization'').
+Directions for accessing the code and operating instances of $ESFNetVis$
+are provided in\footnote{Central repository for Versinus resources such as implementations, public instances of interactive interfaces, and videos: https://github.com/ttm/versinus}.
 
 \begin{figure}[!h]\centering
-\includegraphics[width=\textwidth]{esfNetVis___}
-  \caption{
-  }\label{esfnetvis}
+\includegraphics[width=0.7\textwidth]{esfNetVis___}
+  \caption{Visual elements for the settings performed before
+  the visual mapping of the dynamic network.
+  The buttons at the top are for keeping, modifying and accessing settings.
+  The widgets in the gray rectangles enables the selection of the network, the first and last messages (transactions) to be considered,
+  and the sectorialization method.
+  The widgets in the white rectangle are $\sigma$, $\eta$, and $\omega$.
+  These settings are compliant with the description available in Section~\ref{sec:des}.
+The ``Erdös'' setting stands for the achievement of the network sectors by comparing the network against an Erdös-Rényi model, mentioned in Section~\ref{sec:pla}.
+  When the red button is clicked, the dynamic network is mapped to the canvas and additional elements, as described in Figure~\ref{env2}.
+  }\label{env1}
 \end{figure}
@@
 \begin{figure}[!h]\centering
-\includegraphics[width=\textwidth]{esfNetVis2___}
-  \caption{
-  }\label{esfnetvis}
+\includegraphics[width=0.7\textwidth]{esfNetVis2b_}
+  \caption{The initial visualization of the dynamic network performed using $ESFNetVis$.
+  In the canvas, the network is mapped using a 3D extension of Versinus, in which both segments oscillates vertically and in depth.
+  The red portions of the links denotes the direction of the link (often achieved using arrow heads).
+  The text area starts with overall information of the collection of snapshots (e.g.
+each snapshot yields a mean degree, the set of snapshots yields a mean of such mean degree and a standard deviation of the mean degree).
+  The timelines hold the fractions of nodes in each sector with red, green and blue for hub, intermediary and peripheral nodes.
+  The red bar is the playback cue, i.e. the snapshot being visualized.
+  The analyst may play and pause with the play button,
+  and rewind with the rewind button.
+  It is evident from the timelines that the expected fraction of nodes in each sector is not preserved throughout the network evolution.
+  This entails that in some snapshots the scale-free distribution of connectivity is degraded.
+  }\label{env2}
 \end{figure}
@@
 \begin{figure}[!h]\centering
-\includegraphics[width=\textwidth]{esfNetVis3}
-  \caption{
-  }\label{esfnetvis}
+\includegraphics[width=0.7\textwidth]{esfNetVis3b}
+  \caption{The layout is slightly rotated in order to focus on the peripheral nodes segment.
+This image depicts one of the interaction possibilities among the canvas and the timelines: when the mouse is over a node, the line of the sector of the node in the snapshot is made black.
+  The overall peripheral node is an intermediary node in this snapshot.
+  (Notice the pointer in ``button-mode'', i.e. visualized as a hand.)
+  Also, the text area in this image shows information about specific nodes, provided on-demand to the user, which hits the key ``i'' when the mouse is over a node for the interface to provide the information.
+  }\label{env3}
 \end{figure}
@@
 \begin{figure}[!h]\centering
-\includegraphics[width=\textwidth]{esfNetVis4}
+\includegraphics[width=0.7\textwidth]{esfNetVis4b}
   \caption{
-  }\label{esfnetvis}
+  The layout is rotated in order to focus on the intermediary segment of the sinusoid.
+  The text area shows information about specific snapshots, given on-demand to the analyst, which hits "S" while the snapshot is being visualized.
+  The nodes in the intermediary sectors, visualized on the canvas,
+  are being highlighted by the user: notice the mouse pointer which is over the intermediary sector timeline and the transparent blue spheres around the nodes.
+  }\label{env4}
 \end{figure}
+
 One basic feature of the method is the constant representation of
@@
 to know how the relative sizes of the sectors vary along time, i.e. along
-the sequence of snapshots.
+the sequence of snapshots, and how often nodes migrate between sectors.
 
-\section{Insights gathered}\label{sec:ins}
-This section describes insights gathered by the authors,
-and should be further verified in subsequent work and using
+\section{Hypotheses gathered}\label{sec:ins}
+This section describes hypotheses collected by the authors, mostly about dynamic network layouts, through the development and usage of Versinus.
+Most of these hypotheses should be further verified in subsequent work and using
 suitable validation methods.
-The legitimacy of the exposition herein given is based on the reasonable aspect of the assertions, and on the facts that the authors developed Versinus, used the method extensively, and the scientific contribution which it made possible~\cite{stab}.
+% The legitimacy of the exposition herein given is based on the reasonable aspect of the assertions, and on the facts that the authors developed Versinus, used the method extensively, and the scientific contribution which it made possible~\cite{stab}.
 
 The visualization of network structures is favored by the placement of nodes in an oscillatory pattern, as it tends to decrease the superposition of nodes and links.
-In fact, this principle was implemented in some of the layouts available within ccNetViz, and it may be useful in other well-known layouts, such as the Hive Plot~\cite{hive}.
+In fact, this principle was implemented in some of the layouts available within ccNetViz, and it may be useful for other well-known layouts, such as for the Hive Plot~\cite{hive}.
 Using a single period for the sine is an arbitrary choice, established because it entailed good visualizations for the networks and tasks at hand during the design of Versinus.
-Nonetheless, it is allowed to vary in more recent implementations in JavaScript\footnote{Pivot repository for Versinus resources such as implementations, public instances of interactive interfaces, and videos: https://github.com/ttm/versinus}: the sine in which the hub and intermediary nodes are positioned may have as many periods as parametrized by the analyst, and the same is feature is available for the segment for the peripheral nodes.
-Another important insight is the pertinence of fixed positions for the observation of evolving networks.
+Nonetheless, it is should be allowed to vary in further implementations.
+Another important insight is the pertinence of fixed positions for the observation of dynamic networks.
 Not only this has the potential to reduce the hairball effect~\cite{hairball},
-but it is most often very difficult to keep track of individual nodes and structures when the nodes are allowed to move in the images while the network evolves, as in force-based layouts.
+but it is most often very difficult to keep track of individual nodes and structures when the nodes are allowed to move in the images while the network evolves, e.g.
+in using a force-based layout which is constantly updated.
 The placement of nodes with respect to their topological features (nodes are ordered by connectivity,
@@
 Visual cues such as glyph width, height, shape and color were found informative for grasping information on individual nodes,
-but results in a cognitive overload when attempting to observe global features,
+but often result in a cognitive overload when attempting to observe global features,
 which motivates simplified settings.
@@
 implementation of cues for relating language incident in the networks to their topology, and
-further visually representing topological features (e.g. the fraction of interaction of the intermediary sector with the hubs and with the periphery).
+further visually representing topological features (e.g. the fraction of the interaction of the intermediary sector that is shared with the hubs and with the periphery).
 The experimentation of Versinus features within other layouts is a potentially
 valuable contribution, e.g. wiggling the segments of a Hive Plot~\cite{hive}.
-We should implement further mechanisms for the detection of structural changes within the network evolution, such as using the persistent homology~\cite{hajij}.
-Most importantly, we aim at completing a Visual Analytics utility to support using Versinus in an interactive environment for enhanced exploratory data analysis,
+We should implement further mechanisms for the detection of structural changes within the network evolution, such as persistent homology~\cite{hajij}.
+Most importantly, we aim at completing a Visual Analytics utility to support using Versinus in an interactive environment for exploratory data analysis,
 and verifying the claims described in Section~\ref{sec:ins} through formal validation techniques.
 Finally, we plan to further analyze data sets of evolving networks~\cite{stab} using Versinus.
+
+\subsubsection*{Acknowledgements}
+The authors acknowledge the financial support of the S\~ao Paulo State Research Foundation (FAPESP grant 2017/05838-3).
+The views expressed do not reflect the official policy or position of FAPESP.
 
